### 2017-03-22 rmark_doc_inla_nmixnb_7.Rmd
Add rmarkdown document with code and text

---
title: Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R
author: Timothy D. Meehan^[National Audubon Society, Boulder, Colorado USA] and H&#229;vard
  Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
date: "March 21, 2017"
output: pdf_document
---

##1. Introduction

Successful management of wildlife species requires accurate estimates of abundance (Yoccoz et al. 2001).  One common method for estimating animal abundance is direct counts (Pollock et al. 2002).  Efforts to obtain accurate abundance estimates via direct counts can be hindered by the secretive nature of many wildlife species.  The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated (Wenger and Freeman 2008, Joseph et al. 2009).

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection (Denes et al. 2015).  One such sampling scheme, termed the "metapopulation design" (Kery and Royle 2010), involves repeat visits in rapid succession to each of multiple study sites in a study area.  If abundance at a site is assumed to be constant across repeat visits, an assumption referred to as "population closure", then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.  This detection probability can, in effect, be used to correct abundance estimates for imperfect detection.

Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture or binomial mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, Kery et al. 2005).  A simple form of an N-mixture model describes observed counts at site *i* during survey *j* as coming from a Binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$. Specifically,

$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim Binomial(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site-level covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, *p* is commonly modeled as $logit(p) = \alpha_0 + \alpha_1 * x_i$, a logit-linear function of site-by-survey covariates.

These estimation approaches can be extended to cover *k* distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species (Kery et al. 2009).  In this case, population closure is assumed across *j* surveys within year *k*, but is relaxed across years (Kery et al. 2009). A simple specification of a multiple year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim Binomial(N_{i,k}, p_{i,k})$.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution (Kery and Royle 2010), zero-inflated Poisson distribution (Wenger and Freeman 2008), or observation-level random intercepts (Kery and Schaub 2011).  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution (Martin et al. 2011), parse different components of detection through the use of unique covariates (O'Donnell et al. 2015), or relax assumptions of population closure (Chandler et al. 2011, Dail and Madsen 2011).  We do not discuss all of these variations here, but refer interested readers to Denes et al. (2015) for a nice overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology.  Under the right conditions, abundance estimates can be considerably more accurate than those that ignore detection error.  However, there are practical issues that sometimes act as barriers to adoption.  Much of the development of N-mixture models, and many of the examples in the wildlife literature, implement models using Bayesian modeling software such as WinBUGS, OpenBUGS, or JAGS (Lunn et al. 2012).  These are extremely powerful and flexible platforms for analyzing Bayesian models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the BUGS modeling syntax.  While there are several outstanding learning resources aimed at teaching this skill (Royle and Dorazio 2008, Kery 2010, Kery and Schaub 2011, Kery and Royle 2015) it is, nonetheless, a considerable commitment.  Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.  Coding, tuning, and troubleshooting BUGS models could be considered as much an art as a science.

There are other tools available for analyzing N-mixture models that alleviate these practical issues.  The unmarked package (Fiske et al. 2011) for R statistical computing software (R Core Team 2016) offers several options for analyzing N-mixture models, with the capacity to accommodate overdispersion and dynamic populations.  The model coding syntax used in unmarked is a simple extension of the standard R syntax.  Models are analyzed using a maximum likelihood approach, so model analysis is often completed in a fraction of the time taken using an MCMC approach.  The familiar model syntax and rapid model evaluation of unmarked has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.  However, it comes at a cost – the loss of the rich inferential framework associated with Bayesian analysis (although this capacity is being added in increments).

Here we discuss analysis of N-mixture models using the R-INLA package (Rue et al. 2013) for R.  The R-INLA package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models (Rue et al. 2009).  Like unmarked, the model syntax used in the R-INLA package is a straightforward extension of the modeling syntax commonly used in R.  Also, like unmarked, the computational cost of analyzing models with R-INLA is relatively low when compared to MCMC approaches.  The R-INLA approach is different from unmarked in that inference about model parameters falls within a Bayesian paradigm.

The purpose of this manuscript is to demonstrate analysis of N-mixture models using the R-INLA package. In the process, we simulate a realistic dataset, and analyze it in a Bayesian context using both JAGS, via the runjags package (Denwood 2016) for R, and the R-INLA package. In each case, we demonstrate how data is prepared, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance.

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
# setup
setwd("~/GitHub/Quantitative_Metrics/INLA_NMix")
library(runjags)
library(INLA)
library(ggplot2)
# multiplot function
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  if (is.null(layout)) {
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    for (i in 1:numPlots) {
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
# plotting theme
theme_acbs <- function (base_size = 11, base_family = "") {
  theme_grey(base_size = base_size, base_family = base_family) %+replace%
    theme(panel.background = element_rect(fill = "white", colour = NA),
          panel.border = element_rect(fill = NA, colour = "grey20"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = rel(0.9), angle = 0),
          axis.text.y = element_text(size = rel(0.9), angle = 0),
          strip.background = element_rect(fill = "grey85", colour = "grey20"),
          legend.key = element_rect(fill = "white", colour = NA), complete = TRUE)
}
```

##2. Data simulation

###2.1 Hypothetical scenario

The data simulated for this analysis was intended to represent a typical wildlife study. To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 75 study sites. At each site, 4 replicate surveys are conducted within 6 weeks, during the peak of the breeding season when birds are likely to be singing. In order to estimate a trend in abundance over time, these repeated surveys are conducted over a 9-year period.

In this scenario, the abundance of the species is thought to vary with two site level covariates (covariates 1 and 2) that represents habitat characteristics at a site and do not change appreciably over time. The detection probability is also believed to vary according to two covariates (covariates 1 and 3). The first covariate for detection, covariate 1, is the same site level covariate 1 that affects abundance, although it has the opposite effect on detection. The other detection covariate, covariate 3, is a site-year variable that could be related to average weather conditions during the observations at a given site and year.

Finally, as is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion is accounted for using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values are: p.b0 = 1.0 (detection intercept), p.b1 = -2.0 (effect of detection covariate 1), p.b3 = 0.5 (effect of detection covariate 3), lam.b0 = 2.0 (abundance intercept), lam.b1 = 2.0 (effect of abundance covariate 1), lam.b2 = -3.0 (effect of abundance covariate 2), lam.b4 = 1.0 (effect of year on abundance), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation are centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

###2.2 Simulation function

We simulated data for this study using the following function.

```{r data function, echo=TRUE, message=FALSE, warning=FALSE}
data4sim <- function(n.sites = 75,   # number study sites
                     n.surveys = 4,  # short term replicates
                     n.years = 9,    # number years, MAKE ODD NUMBER
                     lam.b0 = 2.0,   # intercept for log lambda
                     lam.b1 = 2.0,   # slope for log lambda, covariate 1
                     lam.b2 = -3.0,  # slope for log lambda, covariate 2
                     lam.b4 = 1.0,   # slope for log lambda, year
                     p.b0 = 1.0,     # intercept for logit p
                     p.b1 = -2.0,    # slope for logit p, covariate 1
                     p.b3 = 1.0,     # slope for logit p covariate 3
                     disp.size = 3.0 # size of the overdisperison
                     ){
  # setup
  if(n.years %% 2 == 0) {n.years <- n.years + 1}    # make years odd
  y <- array(dim = c(n.sites, n.surveys, n.years))  # array for counts
  N.tr <- array(dim = c(n.sites, n.years))          # array for true abund
  # abundance covariates
  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), scale = F)),
                   dim = c(n.sites, n.years)) # site-level covariate 1
  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), scale = F)),
                   dim = c(n.sites, n.years)) # site-level covariate 2
  yrs <- 1:n.years; yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years)) # std years
  # fill abundance array
  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)	# true lam
  for(i in 1:n.sites){
    for(k in 1:n.years){
    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)   # true N
  }}
  # detection covariates
  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
  x.p.3 <- array(as.numeric(scale(rnorm(n = n.sites * n.years, 0, 0.25),
                                  scale = F)), dim = c(n.sites, n.years))
  # expand x.p.3
  out1 <- c()
  for(k in 1:n.years){
   chunk1 <- x.p.3[,k]
   chunk2 <- rep(chunk1, n.surveys)
   out1 <- c(out1, chunk2)
  }
  x.p.3.arr <- array(out1, dim = c(n.sites, n.surveys, n.years))
  # fill count array
  p.tr <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3.arr) # true p
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y[i, j, k] <- rbinom(1, size = N.tr[i, k], prob = p.tr[i, j, k])
  }}}
  # return data
  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
              x.p.1 = x.p.1[, 1, 1], x.p.3 = x.p.3,
              x.lam.1 = x.lam.1[ ,1], x.lam.2 = x.lam.2[ ,1], yr = yr[1, ], 
              y = y, lam.tr = lam.tr, N.tr = N.tr))
} #end

sim.data <- data4sim()
```

##3.0  Data analysis using JAGS

###3.1 JAGS model specification

We first analyze the simulated data using JAGS. In defining the model, we specify a negative binomial model for the abundance component, and use vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and p.

```{r jags model string, echo=TRUE, message=FALSE, warning=FALSE}
jags.model.string <- "
model {

  # priors
  intP ~ dnorm(0, 0.01)       # detection intercept
  bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
  bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
  intLam ~ dnorm(0, 0.01)     # lambda intercept
  bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
  bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
  bYr ~ dnorm(0, 0.01)        # year effect
  overDisEst ~ dunif(0, 5)    # overdispersion size

  # abundance component
  for (k in 1:nYears){
    for (i in 1:nSites){
      N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
      prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
      log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
                          (bYr * yr[k])

  # detection component
      for (j in 1:nSurveys){
        y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
        p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
        lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
      } # close j loop
    } # close i loop
  } # close k loop

} # close model loop
"
```

###3.2 JAGS parameters, dataset, and initial values

Next, we define the parameters to be monitored during the MCMC runs, bundle the data for JAGS, and create a function for drawing random initial values for the model parameters.  The initial values for abundance are made to avoid values of “NA” and zero, as these would cause computational problems.

```{r params data inits, echo=TRUE, message=FALSE, warning=FALSE}
# parameters to monitor
params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
            "bYr", "overDisEst")
# jags data
jags.data <- list(y = sim.data$y, x.lam.1 = sim.data$x.lam.1, 
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr, 
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
# initial values
N.init <- sim.data$y # initial count values
N.init[is.na(N.init)] <- 1 # clean up NA's
N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01), 
                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01), 
                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01), 
                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01), 
                     overDisEst = runif(1, 0.5, 2.5))
# view data
str(jags.data, digits.d = 1)
```

###3.3 Run JAGS model via runjags and view results

Finally, we set the run parameters and start the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (Gelman-Rubin statistics < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).  Note that the recommended number of effective samples for particularly robust inference is closer to 6000 (Gong and Flegal 2016).  So MCMC processing times reported here could be considered shorter than necessary.

```{r run jags, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
# set run parameters
nc <- 3; na <- 100; nb <- 100; ni <- 500; nt <- 1
# run jags
out.jags <- run.jags(model = jags.model.string, data = jags.data,
                     monitor = params, n.chains = nc, inits = inits,
                     burnin = nb, adapt = na, sample = ni, thin = nt,
                     modules = "glm on", method = "parallel")
```
```{r view jags results}
# view results
round(as.data.frame(summary(out.jags)[,c(2, 1, 3)]), 2)
```

Mean parameter estimates from the JAGS model are reasonably close to, and not significantly different from, the input values used to generate the data: detection intercept = 0.97 versus 1.0, effect of detection covariate 1 = -1.86 versus -2.0, effect of detection covariate 3  = 1.11 versus 1.00, abundance intercept = 2.04 versus 2.0, effect of abundance covariate 1 = 2.09 versus 2.0, effect of abundance covariate 2 = -2.93 versus -3.0, effect of year on abundance = 0.97 versus 1.0, and size of the overdispersion parameter = 2.90 versus 3.0.  The full marginal posterior distributions for model parameters are shown in Figure 1.  The potential scale reduction factor for all variables was < 1.05, and the effective sample size for all variables was > 1640.  The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 20 minutes.

##4.0 Analysis using R-INLA

###4.1 Prepare data for R-INLA

Next, we prepare data for R-INLA, which involved slight modification of the data in the sim.data object, which produced data formatted for JAGS. The object counts.and.count.covs is an R-INLA object that includes an $i*k$ row by j column matrix of counts. Next comes a value of 1, which is turned into an $i*k$ length vector of ones to specify that the model has a global intercept for lambda. Finally, there are two $i*k$ length vector of values for the two static, site covariates of abundance, where the vector of values for i sites is stacked k times.
In addition to the counts.and.count.covs object, we also define x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $i*k$ length vector corresponding with x.p.3.

```{r format data for inla, echo=TRUE, message=FALSE, warning=FALSE}
# format count data
y.inla <- sim.data$y[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y[ , , i]
  y.inla <- rbind(y.inla, y.chunk)
}
# format covariates
x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.inla <- c(sim.data$x.p.3)
# make inla.mdata object
counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, x.lam.2.inla, yr.inla)
str(counts.and.count.covs, digits.d = 2)
```

###4.2 Analyze model using R-INLA and view results

The call to the inla() function includes several components.  First is the model statement.  On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for p and the two covariates for p.  The second argument describes the data, provided here as a list that corresponds with the model formula.  Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture for the count component of the model.  The fourth and fifth arguments specify the priors for the two model components, which are similar to those in the JAGS model.

```{r run inla model, echo=TRUE, message=FALSE, warning=FALSE}
# run inla model
out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)))))
# view detection parameter estimates
round(out.inla$summary.fixed[,c(4, 3, 5)], 2)
# view abundance parameter estimates
round(out.inla$summary.hyperpar[,c(4, 3, 5)], 2)
```

Mean parameter estimates from the R-INLA model are also close to, and not significantly different from, input values: detection intercept = 0.97 versus 1.0, effect of detection covariate 1 = -1.86 versus -2.0, effect of detection covariate 3  = 1.11 versus 1.0, abundance intercept = 2.04 versus 2.0, effect of abundance covariate 1 = 2.09 versus 2.0, effect of abundance covariate 2 = -2.93 versus -3.0, effect of year on abundance = 0.97 versus 1.0, and size of the overdispersion parameter = 1 / 0.34 = 2.94 versus 3.0.  Full marginal posterior distributions from R-INLA are also shown in Figure 1.  The analysis took approximately 6 seconds.

```{r make plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1. Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines)."}
# get output from jags
jags.mcmc <- combine.mcmc(out.jags, thin=3, return.samples=5000)
jags.df <- as.data.frame(jags.mcmc)

# make density for intLam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[1] for NMix observations`)
d1$x <- exp(d1$x)
d2 <- as.data.frame(density(jags.df$intLam)[c(1,2)])
d2$x <- exp(d2$x)
p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab("Density") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1)

# make density for x.1.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[2] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1)

# make density for x.2.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[3] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov2Lam)[c(1,2)])
p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=-3, col="gray20", lty=1)

# make density for yr.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[4] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bYr)[c(1,2)])
p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)

# make density for intP
d1 <- as.data.frame(out.inla$marginals.fixed$`(Intercept)`)
d1$x <- plogis(d1$x)
d2 <- as.data.frame(density(jags.df$intP)[c(1,2)])
d2$x <- plogis(d2$x)
p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)

# make density for x.1.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.1.inla)
d2 <- as.data.frame(density(jags.df$bCov1P)[c(1,2)])
p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)

# make density for x.3.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.3.inla)
d2 <- as.data.frame(density(jags.df$bCov3P)[c(1,2)])
p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab("Density") + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)

# make density for overdisp
d1 <- as.data.frame(out.inla$marginals.hyperpar$`overdispersion for NMix observations`)
d1$x <- 1 / d1$x; d1$y <- d1$y / 8
d2 <- as.data.frame(density(jags.df$overDisEst)[c(1,2)])
p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab("Density") + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)

# plot all
multiplot(p1,p2,p3,p4,p5,p6,p7,p8,cols=3)
```

##5.0 Discussion

The purpose of this study was to (1) demonstrate the use of the R-INLA package to analyze N-mixture models used by wildlife biologists and (2) compare performance of R-INLA to another common approach, JAGS via the runjags package, which uses MCMC methods.

In demonstrating the use of R-INLA, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.  Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for lambda are specified through an R-INLA object included on the left side of the formula.

Regarding performance, both R-INLA and JAGS successfully extracted simulation input values.  Figure 1 depicts marginal posterior distributions produced by JAGS and R-INLA.   These posteriors derive from data resulting from one random manifestation of the input values.  Thus we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.  However, we do expect the input values to fall somewhere within the posterior distributions, which is what occurs here.  Figure 1 shows that, for similarly specified models, R-INLA (dashed black lines) and JAGS (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.  Where R-INLA and JAGS differed was in computing time.  In this example, MCMC took 1,215 seconds to produce posteriors while INLA took 6 seconds – a 200-fold difference.  This was the case despite the fact that the JAGS model was run in parallel with each of three MCMC chains simulated on a separate virtual core.  If parallel computing had not been used, processing the JAGS model would have taken approximately twice as long.  If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have been tripled again.

There are also other differences between the two approaches that, in some cases, will favor the use of JAGS and other BUGS oriented approaches, despite their long processing times.  BUGS based approaches allow users ultimate flexibility in specifying models, whereas, when using R-INLA, only a subset of possible models can be specified.  For example, when using R-INLA, random effects cannot be specified for $\lambda$.  However, it is possible to specify a wide variety of random effects for p, including simple exchangeable intercepts as well as spatially- and temporally-structured random effects.  As another example, when using R-INLA, covariates for $\lambda$ and p need to have the same degree of granularity, e.g., it is not possible to have site-year covariates for lambda and site-survey-year covariates for p.  This limitation might become important if survey level covariates are not somewhat controlled by the sampling design.  In this case, using the means of survey-level covariates per site and year will, at least, allows users to correct for any systematic differences in site-survey-year covariates across sites that might confound estimates of other site-level effects.

##6.0 Appendix A

###6.1 Likelihood

Would you like to add stuff here about the likelihood functions and the iterative procedure for solving them?

##7.0 References

Chandler, R. B., J. A. Royle, and D. I. King. 2011. Inference about density and temporary emigration in unmarked populations. Ecology 92:1429–1435.

Dail, D., and L. Madsen. 2011. Models for estimating abundance from repeated counts of an open metapopulation. Biometrics 67:577–587.

Denes, F. V., L. F. Silveira, and S. R. Beissinger. 2015. Estimating abundance of unmarked animal populations: accounting for imperfect detection and other sources of zero inflation. Methods in Ecology and Evolution 6:543–556.

Denwood, M. J. 2016. runjags: An R package providing interface utilities, model templates, parallel computing methods and additional distributions for MCMC models in JAGS. Journal of Statistical Software 71.

Dodd, C. K., and R. M. Dorazio. 2004. Using counts to simultaneously estimate abundance and detection probabilities in a salamander community. Herpetologica 60:468–478.

Fiske, I., R. Chandler, and others. 2011. unmarked: An R package for fitting hierarchical models of wildlife occurrence and abundance. Journal of Statistical Software 43:1–23.

Gong, L., and J. M. Flegal. 2016. A practical sequential stopping rule for high-dimensional Markov chain Monte Carlo. Journal of Computational and Graphical Statistics 25:684–700.

Joseph, L. N., C. Elkin, T. G. Martin, and H. P. Possingham. 2009. Modeling abundance using N-mixture models: the importance of considering ecological mechanisms. Ecological Applications 19:631–642.

Kery, M. 2010. Introduction to WinBUGS for ecologists: Bayesian approach to regression, ANOVA, mixed models and related analyses. Academic Press.

Kery, M., R. M. Dorazio, L. Soldaat, A. Van Strien, A. Zuiderwijk, and J. A. Royle. 2009. Trend estimation in populations with imperfect detection. Journal of Applied Ecology 46:1163–1172.

Kery, M., and J. A. Royle. 2010. Hierarchical modelling and estimation of abundance and population trends in metapopulation designs. Journal of Animal Ecology 79:453–461.

Kery, M., and J. A. Royle. 2015. Applied hierarchical modeling in ecology: analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models. Academic Press.

Kery, M., J. A. Royle, and H. Schmid. 2005. Modeling avian abundance from replicated counts using binomial mixture models. Ecological applications 15:1450–1461.

Kery, M., and M. Schaub. 2011. Bayesian population analysis using WinBUGS: a hierarchical perspective. Academic Press.

Lunn, D., C. Jackson, N. Best, A. Thomas, and D. Spiegelhalter. 2012. The BUGS book: a practical introduction to Bayesian analysis. CRC press, Boca Raton, Florida.

Martin, J., J. A. Royle, D. I. Mackenzie, H. H. Edwards, M. Kéry, and B. Gardner. 2011. Accounting for non-independent detection when estimating abundance of organisms with a Bayesian approach. Methods in Ecology and Evolution 2:595–601.

O'Donnell, K. M., F. R. Thompson III, and R. D. Semlitsch. 2015. Partitioning detectability components in populations subject to within-season temporary emigration using binomial mixture models. PLoS ONE 10:e0117216.

Pollock, K. H., J. D. Nichols, T. R. Simons, G. L. Farnsworth, L. L. Bailey, and J. R. Sauer. 2002. Large scale wildlife monitoring studies: statistical methods for design and analysis. Environmetrics 13:105–119.

R Core Team. 2016. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.

Royle, J. A. 2004. N-mixture models for estimating population size from spatially replicated counts. Biometrics 60:108–115.

Royle, J. A., and R. M. Dorazio. 2008. Hierarchical modeling and inference in ecology: the analysis of data from populations, metapopulations and communities. Academic Press.

Royle, J. A., and J. D. Nichols. 2003. Estimating abundance from repeated presence–absence data or point counts. Ecology 84:777–790.

Rue, H., S. Martino, and N. Chopin. 2009. Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. Journal of the Royal Statistical Society: Series B (Statistical Methodology) 71:319–392.

Rue, H., S. Martino, F. Lindgren, D. Simpson, and A. Riebler. 2013. R-INLA: approximate Bayesian inference using integrated nested Laplace approximations. Trondheim, Norway.

Wenger, S. J., and M. C. Freeman. 2008. Estimating species occurrence, abundance, and detection probability using zero-inflated distributions. Ecology 89:2953–2959.

Yoccoz, N. G., J. D. Nichols, and T. Boulinier. 2001. Monitoring of biological diversity in space and time. Trends in Ecology & Evolution 16:446–453.

### 2017-04-01 rmark_doc_inla_nmixnb_8.Rmd
Update nmix paper
Added new section on survey level covariate for p and new section using
actual published data.  Still have a ways to go to clean up new
sections.

---
title: Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R
author: Timothy D. Meehan^[National Audubon Society, Boulder, Colorado USA], Nicole L. Michel^[National Audubon Society, San Francisco, California USA], and Haavard Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
date: "March 31, 2017"
output: pdf_document
---

##1. Introduction

###1.1 Background

Successful management of wildlife species requires accurate estimates of abundance (Yoccoz et al. 2001).  One common method for estimating animal abundance is direct counts (Pollock et al. 2002).  Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by methodological factors such as observer expertise, weather, and habitat structure.  The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated (Wenger and Freeman 2008, Joseph et al. 2009).

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection (Denes et al. 2015).  One such sampling scheme, termed the "metapopulation design" (Kery and Royle 2010), involves repeat visits in rapid succession to each of multiple study sites in a study area.  If, during repeat visits, the population is assumed to be closed (no immigration , emmigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.  This detection probability can, in effect, be used to correct abundance estimates for imperfect detection.

Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture or binomial mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, Kery et al. 2005).  A simple form of an N-mixture model describes observed counts *Y* at site *i* during survey *j* as coming from a Binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$. Specifically,

$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim Binomial(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site-level covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, *p* is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-by-survey covariates.

These estimation approaches can be extended to cover *K* distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species (Kery et al. 2009).  In this case, population closure is assumed across *j* surveys within year *k*, but is relaxed across years (Kery et al. 2009). A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim Binomial(N_{i,k}, p_{i,k})$.  Like the single year specification, $\lambda$ is commonly modelled using site and site-by-year covariates, and *p* using site-by-survey-by-year covariates.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution (Kery and Royle 2010), zero-inflated Poisson distribution (Wenger and Freeman 2008), or observation-level random intercepts (Kery and Schaub 2011).  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution (Martin et al. 2011), parse different components of detection through the use of unique covariates (O'Donnell et al. 2015), or relax assumptions of population closure (Chandler et al. 2011, Dail and Madsen 2011).  We do not discuss all of these variations here, but refer interested readers to Denes et al. (2015) for a nice overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology.  Under the right conditions, abundance estimates can be considerably more accurate than those that ignore detection error.  However, there are practical issues that sometimes act as barriers to adoption.  Much of the development of N-mixture models, and many of the examples in the wildlife literature, implement models using Bayesian modeling software such as WinBUGS, OpenBUGS, or JAGS (Lunn et al. 2012).  These are extremely powerful and flexible platforms for analyzing Bayesian models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the BUGS modeling syntax.  While there are several outstanding learning resources aimed at teaching this skill (Royle and Dorazio 2008, Kery 2010, Kery and Schaub 2011, Kery and Royle 2015) it is, nonetheless, a considerable commitment.  Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.  Coding, tuning, and troubleshooting BUGS models could be considered as much an art as a science.

There are other tools available for analyzing N-mixture models that alleviate these practical issues.  The unmarked package (Fiske et al. 2011) for R statistical computing software (R Core Team 2016) offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.  The model coding syntax used in unmarked is a simple extension of the standard R syntax.  Models are analyzed using a maximum likelihood approach, so model analysis is often completed in a fraction of the time taken using an MCMC approach.  The familiar model syntax and rapid model evaluation of unmarked has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.  However, it comes at a cost – the loss of the intuitive inferential framework associated with Bayesian analysis (although this capacity is being added in increments).

Here we discuss analysis of N-mixture models using the R-INLA package (Rue et al. 2013) for R.  The R-INLA package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models (Rue et al. 2009). INLA was developed specifically to allow estimation of posteriors in a fraction of the time taken by MCMC.  Like unmarked, the model syntax used in the R-INLA package is a straightforward extension of the modeling syntax commonly used in R.  Also, like unmarked, the computational cost of analyzing models with R-INLA is relatively low.  The R-INLA approach is different from unmarked in that inference about model parameters falls within a Bayesian paradigm.

###1.2 Objectives

The purpose of this manuscript is to demonstrate analysis of N-mixture models using the R-INLA package. In the process, we explore simulated and actual count datasets, and analyze them using JAGS, via the runjags package (Denwood 2016) for R, the unmarked package, and the R-INLA package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We will also explore a limitation associated with the INLA approach related to model specification.  Specifically, while it is possible to specify survey-level covariates for detection using JAGS and unmarked, this is not possible using INLA.  Rather, survey-level covariates of detection must be averaged to the site or site-by-year level.  Using an averaged detection covariate does allow accounting for site-level differences in survey conditions should they occur.  However, in the process of averaging, important information related to detection is discarded, which could lead to biased abundance estimates under certain conditions. Most of the code used to conduct analyses is shown in body of this manuscript. However, some code related to generating tables and figures is not shown, for brevity, although it can be accessed via https://github.com/tmeeha/inlaNMix.

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
# setup
setwd("~/GitHub/Quantitative_Metrics/INLA_NMix")
library(runjags)
library(INLA)
library(unmarked)
library(ggplot2)
library(knitr)
knitr::opts_chunk$set(cache=TRUE)
# multiplot function
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  if (is.null(layout)) {
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  if (numPlots==1) {
    print(plots[[1]])
  } else {
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    for (i in 1:numPlots) {
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
# plotting theme
theme_acbs <- function (base_size = 11, base_family = "") {
  theme_grey(base_size = base_size, base_family = base_family) %+replace%
    theme(panel.background = element_rect(fill = "white", colour = NA),
          panel.border = element_rect(fill = NA, colour = "grey20"),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = rel(0.9), angle = 0),
          axis.text.y = element_text(size = rel(0.9), angle = 0),
          strip.background = element_rect(fill = "grey85", colour = "grey20"),
          legend.key = element_rect(fill = "white", colour = NA), complete = TRUE)
}
```

###1.3 Simulated data

The data simulated for this analysis was intended to represent a typical wildlife study. To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites. At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season when birds are likely to be singing. In order to estimate a trend in abundance over time, these repeated surveys are conducted over a 9-year period.

In this scenario, the abundance of the species is thought to vary with two site-level covariates (covariates 1 and 2) that represents habitat characteristics at a site and do not change appreciably over time. The detection probability is also believed to vary according to two covariates (covariates 1 and 3). The first covariate for detection, covariate 1, is the same site-level covariate 1 that affects abundance, although it has the opposite effect on detection. The other detection covariate, covariate 3, is a site-survey-year variable that could be related to weather conditions during a given count.

Finally, as is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion is generated using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values are: p.b0 = 1.0 (detection intercept), p.b1 = -2.0 (effect of detection covariate 1), p.b3 = 1.0 (effect of detection covariate 3), lam.b0 = 2.0 (abundance intercept), lam.b1 = 2.0 (effect of abundance covariate 1), lam.b2 = -3.0 (effect of abundance covariate 2), lam.b4 = 1.0 (effect of year on abundance), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation are centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

We simulated data for this study using the following function.  Note that the function produces two versions of one detection covariate, x.p.3 and x.p.3.arr, and two versions of the count matrix, y1 and y2.  x.p.3 is the site-survey-year variable described above.  It is used to generate y2, which is used in excersize 2, below.  x.p.3.mean is derived from x.p.3, where values are unique to site and year, but averaged over surveys.  It is used to generate y1, which is used here, in exercize 1.

```{r data function, echo=TRUE, message=FALSE, warning=FALSE}
data4sim <- function(n.sites = 72,   # number study sites
                     n.surveys = 3,  # short term replicates
                     n.years = 9,    # number years, MAKE ODD NUMBER
                     lam.b0 = 2.0,   # intercept for log lambda
                     lam.b1 = 2.0,   # slope for log lambda, covariate 1
                     lam.b2 = -3.0,  # slope for log lambda, covariate 2
                     lam.b4 = 1.0,   # slope for log lambda, year
                     p.b0 = 1.0,     # intercept for logit p
                     p.b1 = -2.0,    # slope for logit p, covariate 1
                     p.b3 = 1.0,     # slope for logit p covariate 3
                     disp.size = 3.0 # size of the overdisperison
                     ){
  # setup
  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
  N.tr <- array(dim = c(n.sites, n.years))           # array for true abund
  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # array for eg1 counts
  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # array for eg2 counts
  # abundance covariate values
  x.lam.1 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years)) # site-level covariate 1
  x.lam.2 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years)) # site-level covariate 1
  yrs <- 1:n.years; yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2  
  yr <- array(rep(yrs, each=n.sites), dim=c(n.sites, n.years)) # std years
  # fill abundance array
  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)	# true lam
  for(i in 1:n.sites){
    for(k in 1:n.years){
    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size) # true N
  }}
  # detection covariate values
  x.p.1 <- array(x.lam.1[,1], dim=c(n.sites,n.surveys,n.years))
  x.p.3 <- array(as.numeric(scale(runif(n=n.sites*n.surveys*n.years, -0.5, 0.5), 
                                  scale=F)), dim=c(n.sites,n.surveys,n.years))
  # average x.p.3 per site-year
  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
  out1 <- c()
  for(k in 1:n.years){
   chunk1 <- x.p.3.mean[,k]
   chunk2 <- rep(chunk1, n.surveys)
   out1 <- c(out1, chunk2)
  }
  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
  # fill count array with site-yr x.p.3
  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr) # true p
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y1[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr1[i,j,k])
  }}}
  # fill count array with site-surv-yr x.p.3
  p.tr2 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3)     # true p
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y2[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr2[i,j,k])
  }}}
  # return data 
  return(list(n.sites=n.sites, n.surveys=n.surveys, n.years=n.years, 
              x.p.1=x.p.1[,1,1], x.p.3=x.p.3, x.p.3.mean=x.p.3.mean, 
              x.p.3.arr=x.p.3.arr, x.lam.1=x.lam.1[,1], 
              x.lam.2=x.lam.2[,1], yr=yr[1,], y1=y1, y2=y2, 
              lam.tr=lam.tr, N.tr=N.tr))
} #end
```

```{r print make data}
# create dataset for exercise 1 and 2
sim.data <- data4sim()
```

###1.4 Real data

In addition to simulated data, we also demonstrate the use of R-INLA using a real dataset in exercise 3.  This dataset comes from a study by Kery et al. (2005) and is publically available as part of the unmarked package.  The dataset includes mallard duck counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors the abundance of breeding birds. In addition to counts, the dataset also includes 2 site-by-survey covariates related to detection (effort and survey date), and 3 site-level covariates related to abundance (route length, elevation, and forest cover).  Full dataset details are given in Kery et al. (2002).

##2. Exercise 1

###2.1 Goals
In Excersize 1, we demonstrate the use of R-INLA and compare use and performance to a simular analyses using JAGS and unmarked.  In this excercise, the forms of JAGS, unmarked, and INLA models match the data generating process.  Specifically, we used x.p.3.mean to generate y1 and analyzed the data using models that use x.p.3.mean as a covariate.  This exercise demonstrates the differences in package use, computation time, and estimation results.

###2.2 JAGS model specification

We first analyze the simulated data using JAGS. In defining the model, we specify a negative binomial model for the abundance component, and use vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and p.

```{r jags model string, echo=TRUE, message=FALSE, warning=FALSE}
jags.model.string <- "
model {

  # priors
  intP ~ dnorm(0, 0.01)       # detection intercept
  bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
  bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
  intLam ~ dnorm(0, 0.01)     # lambda intercept
  bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
  bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
  bYr ~ dnorm(0, 0.01)        # year effect
  overDisEst ~ dunif(0, 5)    # overdispersion size

  # abundance component
  for (k in 1:nYears){
    for (i in 1:nSites){
      N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
      prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
      log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
                          (bYr * yr[k])

  # detection component
      for (j in 1:nSurveys){
        y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
        p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
        lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
      } # close j loop
    } # close i loop
  } # close k loop

} # close model loop
"
```

###2.3 JAGS parameters, dataset, and initial values

Next, we define the parameters to be monitored during the MCMC runs, bundle the data for JAGS, and create a function for drawing random initial values for the model parameters.  The initial values for abundance are made to avoid values of “NA” and zero, as these would cause computational problems.

```{r params data inits, echo=TRUE, message=FALSE, warning=FALSE}
# parameters to monitor
params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
            "bYr", "overDisEst")
# jags data
jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1, 
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr, 
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
# initial values
N.init <- sim.data$y1 # initial count values
N.init[is.na(N.init)] <- 1 # clean up NA's
N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01), 
                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01), 
                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01), 
                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01), 
                     overDisEst = runif(1, 0.5, 2.5))
```

###2.4 Run JAGS model via runjags and view results

Finally, we set the run parameters and start the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (Gelman-Rubin statistics < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).  Note that the recommended number of effective samples for particularly robust inference is closer to 6000 (Gong and Flegal 2016).  So MCMC processing times reported here could be considered shorter than necessary.

```{r run jags, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
# set run parameters
nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10
# run jags
ptm <- proc.time()
out.jags <- run.jags(model = jags.model.string, data = jags.data,
                     monitor = params, n.chains = nc, inits = inits,
                     burnin = nb, adapt = na, sample = ni, thin = nt,
                     modules = "glm on", method = "parallel")
```

```{r print jags out 1}
# view summary
summary(out.jags)
# computing time
round(jags.time.1 <- proc.time() - ptm, 2)[3]
```

Mean parameter estimates from the JAGS model are reasonably close to, and not significantly different from, the input values used to generate the data (Table 1).  The full marginal posterior distributions for model parameters are shown in Figure 1.  The potential scale reduction factor for all variables was < 1.05, and the effective sample size for all variables was > 1640.  The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 20 minutes.

###2.5 Prepare data for unmarked

Next, we prepare data for the unmarked package, which involved slight modification of the data in the sim.data object.  Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by *J* column matrix.  Each static, site-level variable was duplicated and stacked *K* times to form a single column vector. A column vector was created to identify each year in the stacked data.  The site-by-year variable, x.p.3, was transformed from 2-dimensional matrix to a single column vector.  Reformatted variables were then assembled in an unmarked data structure called an unmarked frame.

```{r format data for unmk, echo=TRUE, message=FALSE, warning=FALSE}
# format count data
y.unmk <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.unmk <- rbind(y.unmk, y.chunk)
}
# format covariates
x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.unmk <- c(sim.data$x.p.3.mean)
site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk, 
                             x.p.1.unmk, x.p.3.unmk)
# make unmarked pcount frame
unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
```

###2.6 Analyze model using unmarked and view results

The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance.  This was followed by an argument identifying the appropriate unmarked frame and the form of the count portion of the mixture model, negative binomial.

```{r run unmk model, echo=TRUE, message=FALSE, warning=FALSE}
# run unmk model
ptm <- proc.time()
out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
                   ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
                    data = unmk.data, mixture = "NB")
```

```{r print unmk out 1}
# view summary
summary(out.unmk)
# computing time
round(unmk.time.1 <- proc.time() - ptm, 2)[3]
```

Maximum liklihood estimates from the unmarked analysis are are also close to, and not significantly different from, input values (Table 1)  The analysis took approximately 120 seconds.

###2.7 Prepare data for R-INLA

Next, we prepare data for R-INLA, which involved slight modification of the data in the sim.data object, which produced data formatted for JAGS. The object counts.and.count.covs is an R-INLA object that includes an $I*K$ row by j column matrix of counts. Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for lambda. Finally, there are two $I*K$ length vector of values for the two static, site covariates of abundance, where the vector of values for i sites is stacked k times.
In addition to the counts.and.count.covs object, we also define x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.

```{r format data for inla, echo=TRUE, message=FALSE, warning=FALSE}
# format count data
y.inla <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.inla <- rbind(y.inla, y.chunk)
}
# format covariates
x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.inla <- c(sim.data$x.p.3.mean)
# make inla.mdata object
counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, x.lam.2.inla, yr.inla)
```

###2.8 Analyze model using R-INLA and view results

The call to the inla() function includes several components.  First is the model statement.  On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for p and the two covariates for p.  The second argument describes the data, provided here as a list that corresponds with the model formula.  Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture for the count component of the model.  The fourth and fifth arguments specify the priors for the two model components, which are similar to those in the JAGS model.

```{r run inla model, echo=TRUE, message=FALSE, warning=FALSE}
# run inla model
ptm <- proc.time()
out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)))))
```

```{r print inla out 1}
# view summary
summary(out.inla)
# computing time
round(unmk.time.1 <- proc.time() - ptm, 2)[3]
```

Mean parameter estimates from the R-INLA model are also close to, and not significantly different from, input values (Table 1)  Full marginal posterior distributions from R-INLA are also shown in Figure 1.  The analysis took approximately 6 seconds.

```{r output table, echo=FALSE}
# get jags out
jags.out.tab <- round(as.data.frame(summary(out.jags)[,c(2, 1, 3)]), 3)
# get unmk out
e1 <- as.matrix(out.unmk@estimates@estimates$det@estimates)
ci1 <- as.matrix(confint(out.unmk, type="det"))
e2 <- as.matrix(out.unmk@estimates@estimates$state@estimates)
ci2 <- as.matrix(confint(out.unmk, type="state"))
e3 <- as.matrix(out.unmk@estimates@estimates$alpha@estimates)
ci3 <- as.matrix(confint(out.unmk, type="alpha"))
e4 <- rbind(e1,e2,e3)
ci4 <- rbind(ci1,ci2,ci3)
unmk.out.tab <- cbind(e4,ci4)
# get inla out
in1 <- out.inla$summary.fixed[,c(4, 3, 5)]
in2 <- out.inla$summary.hyperpar[,c(4, 3, 5)][-5,]
in3 <- 1/out.inla$summary.hyperpar[,c(4, 3, 5)][5,]
in3 <- in3[,c(1,3,2)]
inla.out.tab <- rbind(in1,in2,in3)
# combine out
all.out.tab <- round(cbind(jags.out.tab, unmk.out.tab, inla.out.tab),2)
colnames(all.out.tab) <- c("JAGS Median","JAGS Lower CrL", "JAGS Upper CrL",
                           "UNMK Estimate","UNMK Lower CL", "UNMK Upper CL",
                           "INLA Median","INLA Lower CrL", "INLA Upper CrL")
rownames(all.out.tab) <- c("P Int", "P Cov 1", "P Cov 3", "Lambda Int", "Lambda Cov 1",
                           "Lambda Cov 2", "Lambda Year", "Overdispersion")
```

```{r print table, echo=F}
print(kable(all.out.tab, align="c", caption="Parameter estimates from JAGS, unmarked, and R-INLA"))
```

###2.9 Notes on Exercise 1

Exercise 1 shows several things.

In demonstrating the use of R-INLA, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.  Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for lambda are specified through an R-INLA object included on the left side of the formula.

Regarding performance, both R-INLA and JAGS successfully extracted simulation input values.  Figure 1 depicts marginal posterior distributions produced by JAGS and R-INLA.   These posteriors derive from data resulting from one random manifestation of the input values.  Thus we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.  However, we do expect the input values to fall somewhere within the posterior distributions, which is what occurs here.  Figure 1 shows that, for similarly specified models, R-INLA (dashed black lines) and JAGS (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.  Where R-INLA and JAGS differed was in computing time.  In this example, MCMC took 1,215 seconds to produce posteriors while INLA took 6 seconds – a 200-fold difference.  This was the case despite the fact that the JAGS model was run in parallel with each of three MCMC chains simulated on a separate virtual core.  If parallel computing had not been used, processing the JAGS model would have taken approximately twice as long.  If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have been tripled again.

```{r make plot 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1. Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines)."}
# get output from jags
jags.mcmc <- combine.mcmc(out.jags, thin=3, return.samples=5000)
jags.df <- as.data.frame(jags.mcmc)
# make density for intLam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[1] for NMix observations`)
d1$x <- exp(d1$x)
d2 <- as.data.frame(density(jags.df$intLam)[c(1,2)])
d2$x <- exp(d2$x)
p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1)
# make density for x.1.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[2] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1)
# make density for x.2.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[3] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bCov2Lam)[c(1,2)])
p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab(" ") + theme_acbs() + geom_vline(xintercept=-3, col="gray20", lty=1)
# make density for yr.lam
d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[4] for NMix observations`)
d2 <- as.data.frame(density(jags.df$bYr)[c(1,2)])
p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
# make density for intP
d1 <- as.data.frame(out.inla$marginals.fixed$`(Intercept)`)
d1$x <- plogis(d1$x)
d2 <- as.data.frame(density(jags.df$intP)[c(1,2)])
d2$x <- plogis(d2$x)
p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)
# make density for x.1.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.1.inla)
d2 <- as.data.frame(density(jags.df$bCov1P)[c(1,2)])
p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)
# make density for x.3.p
d1 <- as.data.frame(out.inla$marginals.fixed$x.p.3.inla)
d2 <- as.data.frame(density(jags.df$bCov3P)[c(1,2)])
p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
# make density for overdisp
d1 <- as.data.frame(out.inla$marginals.hyperpar$`overdispersion for NMix observations`)
d1$x <- 1 / d1$x; d1$y <- d1$y / 8
d2 <- as.data.frame(density(jags.df$overDisEst)[c(1,2)])
p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)
# plot all
multiplot(p1,p2,p3,p4,p5,p6,p7,p8,cols=3)
```

##3. Example 2, Simulated Data

###3.1 Goals
In Example 2, we show the consequences of not being able to specify a survey level covariate.

There are also other differences between the two approaches that, in some cases, will favor the use of JAGS and other BUGS oriented approaches, despite their long processing times.  BUGS based approaches allow users ultimate flexibility in specifying models, whereas, when using R-INLA, only a subset of possible models can be specified.  For example, when using R-INLA, random effects cannot be specified for $\lambda$.  However, it is possible to specify a wide variety of random effects for p, including simple exchangeable intercepts as well as spatially- and temporally-structured random effects.  As another example, when using R-INLA, covariates for $\lambda$ and p need to have the same degree of granularity, e.g., it is not possible to have site-year covariates for lambda and site-survey-year covariates for p.  This limitation might become important if survey level covariates are not somewhat controlled by the sampling design.  In this case, using the means of survey-level covariates per site and year will, at least, allows users to correct for any systematic differences in site-survey-year covariates across sites that might confound estimates of other site-level effects.

```{r simulation function for ssy cov, echo=FALSE}
# simulation function
sim.fun <- function(){
  # same inputs as before for most parameters
  n.sites = 72; n.surveys = 3; n.years = 9 
  lam.b0 = 2.0; lam.b1 = 2.0; lam.b2 = -3.0; lam.b4 = 1.0
  p.b0 = 1.0; p.b1 = -2.0
  disp.size = 3.0
  # now vary the effect size for detection covariate 3
  p.b3 = runif(1, -3.0, 3.0)
  # keep track of input values for later analysis
  real.vals <- c(p.b0, p.b1, p.b3, lam.b0, lam.b1, lam.b2, lam.b4, disp.size)
  # set jags run parameters
  nc <- 3; na<- 500; nb <- 100; ni <- 1000; nt <- 1
  # parameters to monitor
  params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
            "bYr", "overDisEst")
  # make data
  sim.data <- data4sim(n.sites = n.sites, n.surveys = n.surveys, 
                       n.years = n.years,
                       lam.b0 = lam.b0, lam.b1 = lam.b1, lam.b2 = lam.b2,
                       lam.b4 = lam.b4,
                       p.b0 = p.b0, p.b1 = p.b1, p.b3 = p.b3,
                       disp.size = disp.size)
  # bundle jags with site-survey-year x.p.3
  jags.data.big <- list(y = sim.data$y2, x.lam.1 = sim.data$x.lam.1, 
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr, 
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
  # initial values for first jags run
  N.init <- sim.data$y2 # initial count values
  N.init[is.na(N.init)] <- 1 # clean up NA's
  N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
  inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01), 
                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01), 
                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01), 
                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01), 
                     overDisEst = runif(1, 0.5, 2.5))
  # new jags model with expanded x.p.3
  jags.model.string.expand <- "
  model {
    # priors
    intP ~ dnorm(0, 0.01)       # detection intercept
    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
    intLam ~ dnorm(0, 0.01)     # lambda intercept
    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
    bYr ~ dnorm(0, 0.01)        # year effect
    overDisEst ~ dunif(0, 5)    # overdispersion size
    # abundance component
    for (k in 1:nYears){
      for (i in 1:nSites){
        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
                            (bYr * yr[k])
    # detection component, note new i, j, k, subscript for x.p.3
        for (j in 1:nSurveys){
          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, j, k])
        } # close j loop
      } # close i loop
    } # close k loop
  } # close model loop
  "
  # call jags for site-survey-year x.p.3
  out.jags.big <- run.jags(model=jags.model.string.expand, data=jags.data.big,
                           monitor=params, n.chains=nc, inits=inits,
                           burnin=nb, adapt=na, sample=ni,
                           thin=nt, modules="glm on", method="parallel")
  # collect results
  bo <- as.data.frame(round(summary(out.jags.big),2))[,c(1:5,9,11)]
  bo$simtype <- "big"; bo$p.b3 <- p.b3; bo$par <- row.names(bo)
  bo$true.vals <- real.vals
  # bundle jags data with site-year x.p.3
  jags.data.small <- list(y = sim.data$y2, x.lam.1 = sim.data$x.lam.1, 
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr, 
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.arr,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
  # call jags for site-year x.p.3
  out.jags.small <- run.jags(model=jags.model.string.expand, data=jags.data.small,
                             monitor=params, n.chains=nc, inits=inits,
                             burnin=nb, adapt=na, sample=ni,
                             thin=nt, modules="glm on", method="parallel")
  so <- as.data.frame(round(summary(out.jags.small),2))[,c(1:5,9,11)]
  so$simtype <- "small"; so$p.b3 <- p.b3; so$par <- row.names(so)
  so$true.vals <- real.vals
  # combine results
  all.out <- rbind(bo,so)
  all.out$diffs <- all.out$Median - all.out$true.vals
  all.out$range <- all.out$Lower95 - all.out$Upper95
  row.names(all.out) <- NULL
  return(all.out)
}
```

```{r run simulation, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
# first round
ptm <- proc.time()
sim.out <- sim.fun()
# subsequent rounds
nsims <- 49
for(w in 2:nsims){
  iter.out <- sim.fun()
  sim.out <- rbind(sim.out, iter.out)
  cat(paste("\n\n\n this is iteration", w, "\n\n\n"))
}
proc.time() - ptm
```

```{r explore sim results, echo=FALSE}
# explore results
par_names <- c(
  'bCov1Lam'="Abundance covariate 1 (2)",
  'bCov2Lam'="Abundance covariate 2 (-3)",
  'bYr'="Abundance covariate year (1)",
  'bCov1P'="Detection covariate 1 (-2)",
  'bCov3P'="Detection covariate 3 (X)",
  'intLam'="Abundance intercept (2)",
  'intP'="Detection intercept (1)",
  'overDisEst'="Overdispersion estimate (3)"
)
to.string <- as_labeller(par_names)
ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=0.5) +
  facet_wrap(~factor(par), scales="free",labeller=to.string) +
  xlab("Slope coefficient for the detection covariate") +
  ylab("Difference between posterior mean and true parameter value") +
  geom_smooth(span=3, se=F, size=0.5) + 
  theme_acbs() + scale_color_manual(values=c("black","gray50")) + 
  theme(strip.background = element_rect(colour = "black", fill = "white"), 
        legend.position="none") + scale_linetype_manual(values=c(1,2))
```

###3.2  Results

This is what happened in this section.

##4. Example 3, Real Data

###4.1 Goals

We wanted to show real data.

###4.1 unmarked

```{r second unmarked analysis}
# get real mallard data from unmarked package
data(mallard)
# format unmarked data 
mallard.y.unmk <- mallard.y
mallard.site.unmk <- mallard.site
mallard.obs.unmk <- mallard.obs
mallardUMF <- unmarkedFramePCount(mallard.y.unmk, siteCovs = mallard.site.unmk,
                                  obsCovs = mallard.obs.unmk)
# run unmarked model
mallard.out.unmk <- pcount(~ ivel+ date + I(date^2) ~ length + elev + forest, 
                           mixture = "NB", mallardUMF, K=30)
```

```{r summary for unmk out 2}
summary(mallard.out.unmk)
```

###4.2 INLA

```{r second inla analysis}
# format inla data
mallard.y.inla <- mallard.y
mallard.length.inla <- mallard.site[,2]
mallard.elev.inla <- mallard.site[,1]
mallard.forest.inla <- mallard.site[,3]
mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T)
mallard.ivel.inla[is.na(mallard.ivel.inla)] <- mean(mallard.ivel.inla, na.rm=T)
mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T)
mallard.date2.inla <- mallard.date.inla^2
# make inla.mdata object
counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, mallard.length.inla, 
                                    mallard.elev.inla, mallard.forest.inla)
# run inla model
mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla + 
                           mallard.date.inla + mallard.date2.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     mallard.ivel.inla = mallard.ivel.inla, 
                     mallard.date.inla = mallard.date.inla,
                     mallard.date2.inla = mallard.date2.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)))))
```

```{r summary for out inla 2}
summary(mallard.out.inla)
```

```{r exercize three plot, echo=F}
# get unmk out
e1 <- as.matrix(mallard.out.unmk@estimates@estimates$det@estimates)
ci1 <- as.matrix(confint(mallard.out.unmk, type="det"))
e2 <- as.matrix(mallard.out.unmk@estimates@estimates$state@estimates)
ci2 <- as.matrix(confint(mallard.out.unmk, type="state"))
e3 <- exp(as.matrix(mallard.out.unmk@estimates@estimates$alpha@estimates))
ci3 <- exp(as.matrix(confint(mallard.out.unmk, type="alpha")))
e4 <- rbind(e1,e2,e3)
ci4 <- rbind(ci1,ci2,ci3)
unmk.out.tab.2 <- cbind(e4,ci4); colnames(unmk.out.tab.2) <- c("X1", "X2", "X3")
# get inla out
in1 <- mallard.out.inla$summary.fixed[,c(4, 3, 5)]
in2 <- mallard.out.inla$summary.hyperpar[,c(4, 3, 5)][-5,]
in3 <- 1/mallard.out.inla$summary.hyperpar[,c(4, 3, 5)][5,]
in3 <- in3[,c(1,3,2)]
inla.out.tab.2 <- rbind(in1,in2,in3); colnames(inla.out.tab.2) <- c("X1", "X2", "X3")
# combine out
all.out.tab.2 <- round(rbind(unmk.out.tab.2, inla.out.tab.2),2)
colnames(all.out.tab.2) <- c("Estimate","Lower", "Upper")
all.out.tab.2$Parameter <- factor(rep(c("P Int", "P Ivel", "P Date", "P Date Sq",
                           "Lambda Int", "Lambda Length", "Lambda Elev", "Lambda Forest", 
                           "Overdispersion"), 2))
all.out.tab.2$Technique <- factor(rep(c("unmarked", "R-INLA"), each=9))
rownames(all.out.tab.2) <- NULL
pd <- position_dodge(0.2) 
ggplot(data=all.out.tab.2, aes(y=Estimate, x=Parameter, col=Technique)) + geom_point(position=pd) + 
  geom_errorbar(aes(ymin = Lower, ymax = Upper), position=pd) + 
  theme_acbs() + scale_color_manual(values=c("black","gray50")) + coord_flip()
```

This is what we found.

##5. Discussion

The purpose of this study was to (1) demonstrate the use of the R-INLA package to analyze N-mixture models used by wildlife biologists and (2) compare performance of R-INLA to another common approach, JAGS via the runjags package, which uses MCMC methods.

##6.0 Appendix A

###6.1 Likelihood

Would you like to add stuff here about the likelihood functions and the iterative procedure for solving them?

##7.0 References

Chandler, R. B., J. A. Royle, and D. I. King. 2011. Inference about density and temporary emigration in unmarked populations. Ecology 92:1429–1435.

Dail, D., and L. Madsen. 2011. Models for estimating abundance from repeated counts of an open metapopulation. Biometrics 67:577–587.

Denes, F. V., L. F. Silveira, and S. R. Beissinger. 2015. Estimating abundance of unmarked animal populations: accounting for imperfect detection and other sources of zero inflation. Methods in Ecology and Evolution 6:543–556.

Denwood, M. J. 2016. runjags: An R package providing interface utilities, model templates, parallel computing methods and additional distributions for MCMC models in JAGS. Journal of Statistical Software 71.

Dodd, C. K., and R. M. Dorazio. 2004. Using counts to simultaneously estimate abundance and detection probabilities in a salamander community. Herpetologica 60:468–478.

Fiske, I., R. Chandler, and others. 2011. unmarked: An R package for fitting hierarchical models of wildlife occurrence and abundance. Journal of Statistical Software 43:1–23.

Gong, L., and J. M. Flegal. 2016. A practical sequential stopping rule for high-dimensional Markov chain Monte Carlo. Journal of Computational and Graphical Statistics 25:684–700.

Joseph, L. N., C. Elkin, T. G. Martin, and H. P. Possingham. 2009. Modeling abundance using N-mixture models: the importance of considering ecological mechanisms. Ecological Applications 19:631–642.

Kery, M. 2010. Introduction to WinBUGS for ecologists: Bayesian approach to regression, ANOVA, mixed models and related analyses. Academic Press.

Kery, M., R. M. Dorazio, L. Soldaat, A. Van Strien, A. Zuiderwijk, and J. A. Royle. 2009. Trend estimation in populations with imperfect detection. Journal of Applied Ecology 46:1163–1172.

Kery, M., and J. A. Royle. 2010. Hierarchical modelling and estimation of abundance and population trends in metapopulation designs. Journal of Animal Ecology 79:453–461.

Kery, M., and J. A. Royle. 2015. Applied hierarchical modeling in ecology: analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models. Academic Press.

Kery, M., J. A. Royle, and H. Schmid. 2005. Modeling avian abundance from replicated counts using binomial mixture models. Ecological applications 15:1450–1461.

Kery, M., and M. Schaub. 2011. Bayesian population analysis using WinBUGS: a hierarchical perspective. Academic Press.

Lunn, D., C. Jackson, N. Best, A. Thomas, and D. Spiegelhalter. 2012. The BUGS book: a practical introduction to Bayesian analysis. CRC press, Boca Raton, Florida.

Martin, J., J. A. Royle, D. I. Mackenzie, H. H. Edwards, M. Kéry, and B. Gardner. 2011. Accounting for non-independent detection when estimating abundance of organisms with a Bayesian approach. Methods in Ecology and Evolution 2:595–601.

O'Donnell, K. M., F. R. Thompson III, and R. D. Semlitsch. 2015. Partitioning detectability components in populations subject to within-season temporary emigration using binomial mixture models. PLoS ONE 10:e0117216.

Pollock, K. H., J. D. Nichols, T. R. Simons, G. L. Farnsworth, L. L. Bailey, and J. R. Sauer. 2002. Large scale wildlife monitoring studies: statistical methods for design and analysis. Environmetrics 13:105–119.

R Core Team. 2016. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.

Royle, J. A. 2004. N-mixture models for estimating population size from spatially replicated counts. Biometrics 60:108–115.

Royle, J. A., and R. M. Dorazio. 2008. Hierarchical modeling and inference in ecology: the analysis of data from populations, metapopulations and communities. Academic Press.

Royle, J. A., and J. D. Nichols. 2003. Estimating abundance from repeated presence–absence data or point counts. Ecology 84:777–790.

Rue, H., S. Martino, and N. Chopin. 2009. Approximate Bayesian inference for latent Gaussian models by using integrated nested Laplace approximations. Journal of the Royal Statistical Society: Series B (Statistical Methodology) 71:319–392.

Rue, H., S. Martino, F. Lindgren, D. Simpson, and A. Riebler. 2013. R-INLA: approximate Bayesian inference using integrated nested Laplace approximations. Trondheim, Norway.

Wenger, S. J., and M. C. Freeman. 2008. Estimating species occurrence, abundance, and detection probability using zero-inflated distributions. Ecology 89:2953–2959.

Yoccoz, N. G., J. D. Nichols, and T. Boulinier. 2001. Monitoring of biological diversity in space and time. Trends in Ecology & Evolution 16:446–453.

### 2017-04-02 rmark_doc_inla_nmixnb_8.Rmd
More updates

@@
 date: "March 31, 2017"
-output: pdf_document
+output: 
+  pdf_document: 
+    fig_caption: yes
 ---
@@
 d2$x <- exp(d2$x)
-p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1)
+col1 <- 4
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- exp(c(d3,d4,d5))
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Lambda") + ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() + geom_vline(xintercept=exp(2), col="gray20", lty=1) +
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.1.lam
@@
 d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
-p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1)
+col1 <- 5
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- c(d3,d4,d5)
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() + geom_vline(xintercept=2, col="gray20", lty=1) +
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.2.lam
@@
 d2 <- as.data.frame(density(jags.df$bCov2Lam)[c(1,2)])
-p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab(" ") + theme_acbs() + geom_vline(xintercept=-3, col="gray20", lty=1)
+col1 <- 6
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- c(d3,d4,d5)
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p3 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Abundance covariate 2")  + ylab(" ") + theme_acbs() + 
+geom_vline(xintercept=-3, col="gray20", lty=1) + 
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for yr.lam
@@
 d2 <- as.data.frame(density(jags.df$bYr)[c(1,2)])
-p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
+col1 <- 7
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- c(d3,d4,d5)
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p4 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Year covariate")  + ylab(NULL) + theme_acbs() + 
+geom_vline(xintercept=1, col="gray20", lty=1) + 
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for intP
@@
 ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=0.5) +
-  facet_wrap(~factor(par), scales="free",labeller=to.string) +
+  facet_wrap(~factor(par), scales="fixed",labeller=to.string) +
   xlab("Slope coefficient for the detection covariate") +

### 2017-04-02 Appendix/appendix.tex
Added the Appendix

\documentclass[a4paper,11pt]{article}
\usepackage{amstext}
\usepackage{amsmath}

\begin{document}

\appendix
\section{Recursive computations of the ``NMix'' likelihood}

The likelihood for the simplest case is
\begin{displaymath}
    \text{Prob}(y) = \sum_{n = y}^{\infty}
    \text{Pois}(n ; \lambda) \;\times\; \text{Binom}(y;  n, p)
\end{displaymath}
where $\text{Poiss}(n; \lambda)$ is the density for the Poisson
distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and
$\text{Binom}(y; n, p)$ is the density for the Binomial distribution
with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
Although the likelhood can be computed directly when replacing the
infinite limit with a finite value, we will demonstrate here that we
can easily evaluate it using a recursive algorithm that is both faster
and more numerical stable. The same idea is also applicable to the
negative Binomial case, and the case where we have replicated
observations of the same ``$n$''. We leave it to the reader to derive
these straight forward extentions.

The key observation is that both the Poisson and the Binomial
distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Pois}(n; \lambda) = \text{Pois}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binom}(y; n, p) = \text{Binom}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Pois}(n ; \lambda) \; \text{Binom}(y;  n, p)
    =
    \text{Pois}(n-1; \lambda) \; \text{Binom}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make
use of this recursive form to express the likelihood with a finite
upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Pois}(n ; \lambda)\;
                      \text{Binom}(y;  n, p) \nonumber\\
                   &=& \text{Pois}(y; \lambda)\; \text{Binom}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 + 
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Pois}(y; \lambda)\; \text{Binom}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then evaluated using the following simple
\texttt{R}-code
\begin{verbatim}
  fac = 1; ff = lambda * (1-p)
  for (i in (n.max - y):1) fac = 1 + fac * ff / i
  log.L = dpois(y, lambda, log=TRUE) +
          dbinom(y, y, p, log=TRUE) + log(fac)
\end{verbatim}
Since this evaluation is recursive in decreasing $n$, we have to chose
the upper limit $n_\text{max}$ up-front, for example as an integer
larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is
small. Note that we are computing \texttt{fac} starting with the
smallest contributions, which are more numerical stable.

\end{document}

% LocalWords:

%%% Local Variables: 
%%% TeX-master: t
%%% End:

### 2017-04-07 rmark_doc_inla_nmixnb_8.Rmd
new draft of manuscript

@@
 Meehan^[National Audubon Society, Boulder, Colorado USA], Nicole L.
-Michel^[National Audubon Society, San Francisco, California USA], and Haavard Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
-date: "March 31, 2017"
+Michel^[National Audubon Society, San Francisco, California USA], and H&#229;vard Rue^[King Abdulla University of Science and Technology, Thuwal SAU]
+date: "April 7, 2017"
+fontsize: 11pt
 output: 
@@
 
-##1. Introduction
+#1. Introduction
 
@@
 2002).
-Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by methodological factors such as observer expertise, weather, and habitat structure.
+Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure.
 The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated (Wenger and Freeman 2008, Joseph et al.
@@
 2015).
-One such sampling scheme, termed the "metapopulation design" (Kery and Royle 2010), involves repeat visits in rapid succession to each of multiple study sites in a study area.
-If, during repeat visits, the population is assumed to be closed (no immigration , emmigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
+One such sampling scheme, termed the "metapopulation design" (K&#233;ry and Royle 2010), involves repeat visits in rapid succession to each of multiple study sites in a study area.
+If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
 This detection probability can, in effect, be used to correct abundance estimates for imperfect detection.
 
-Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture or binomial mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, Kery et al.
+Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture or binomial mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, K&#233;ry et al.
 2005).
-A simple form of an N-mixture model describes observed counts *Y* at site *i* during survey *j* as coming from a Binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$.
+A simple form of an N-mixture model describes observed counts *Y* at site *i* during survey *j* as coming from a binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$.
 Specifically,
 
-$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim Binomial(N_i, p).$$
-
-$\lambda$ is commonly modeled as a log-linear function of site-level covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$.
-Similarly, *p* is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-by-survey covariates.
-
-These estimation approaches can be extended to cover *K* distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species (Kery et al.
+$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim binomial(N_i, p).$$
+
+$\lambda$ is commonly modeled as a log-linear function of site covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$.
+Similarly, *p* is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.
+
+These estimation approaches can be extended to cover *K* distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species (K&#233;ry et al.
 2009).
-In this case, population closure is assumed across *j* surveys within year *k*, but is relaxed across years (Kery et al.
+In this case, population closure is assumed across *J* surveys within year *k*, but is relaxed across years (K&#233;ry et al.
 2009).
-A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim Binomial(N_{i,k}, p_{i,k})$.
-Like the single year specification, $\lambda$ is commonly modelled using site and site-by-year covariates, and *p* using site-by-survey-by-year covariates.
-There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution (Kery and Royle 2010), zero-inflated Poisson distribution (Wenger and Freeman 2008), or observation-level random intercepts (Kery and Schaub 2011).
+A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim binomial(N_{i,k}, p_{i,k})$.
+Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and *p* using site-survey-year covariates.
+There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution (K&#233;ry and Royle 2010), zero-inflated Poisson distribution (Wenger and Freeman 2008), or observation-level random intercepts (K&#233;ry and Schaub 2011).
 Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution (Martin et al.
@@
 The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology.
-Under the right conditions, abundance estimates can be considerably more accurate than those that ignore detection error.
 However, there are practical issues that sometimes act as barriers to adoption.
@@
 First, many wildlife biologists are not accustomed to coding statistical models using the BUGS modeling syntax.
-While there are several outstanding learning resources aimed at teaching this skill (Royle and Dorazio 2008, Kery 2010, Kery and Schaub 2011, Kery and Royle 2015) it is, nonetheless, a considerable commitment.
+While there are several outstanding resources aimed at teaching this skill (Royle and Dorazio 2008, K&#233;ry 2010, K&#233;ry and Schaub 2011, K&#233;ry and Royle 2015) it is, nonetheless, a considerable commitment.
 Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.
-Coding, tuning, and troubleshooting BUGS models could be considered as much an art as a science.
 
@@
 The model coding syntax used in unmarked is a simple extension of the standard R syntax.
-Models are analyzed using a maximum likelihood approach, so model analysis is often completed in a fraction of the time taken using an MCMC approach.
+Models are analyzed using a Maximum Likelihood (ML) approach, so model analysis is often completed in a fraction of the time taken using an MCMC approach.
 The familiar model syntax and rapid model evaluation of unmarked has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.
-However, it comes at a cost – the loss of the intuitive inferential framework associated with Bayesian analysis (although this capacity is being added in increments).
+However, it comes at a cost – the loss of the intuitive inferential framework associated with Bayesian analysis.
 
@@
 2009).
-INLA was developed specifically to allow estimation of posteriors in a fraction of the time taken by MCMC.
+INLA was developed to allow estimation of posteriors in a fraction of the time taken by MCMC.
 Like unmarked, the model syntax used in the R-INLA package is a straightforward extension of the modeling syntax commonly used in R.
@@
 
-###1.2 Objectives
+###1.2 Overall objectives
 
 The purpose of this manuscript is to demonstrate analysis of N-mixture models using the R-INLA package.
-In the process, we explore simulated and actual count datasets, and analyze them using JAGS, via the runjags package (Denwood 2016) for R, the unmarked package, and the R-INLA package.
+In the process, we employ simulated and real count datasets, and analyze them using JAGS, via the runjags package (Denwood 2016) for R, the unmarked package (Fiske et al.
+2011), and the R-INLA package (Rue et al.
+2013).
 In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance.
-We will also explore a limitation associated with the INLA approach related to model specification.
-Specifically, while it is possible to specify survey-level covariates for detection using JAGS and unmarked, this is not possible using INLA.
-Rather, survey-level covariates of detection must be averaged to the site or site-by-year level.
+We also explore a limitation of the R-INLA approach, related to model specification.
+Specifically, while it is possible to specify survey level covariates for detection using JAGS and unmarked, this is not possible using INLA.
+Rather, survey level covariates of detection must be averaged to the site or site-year level.
 Using an averaged detection covariate does allow accounting for site-level differences in survey conditions should they occur.
-However, in the process of averaging, important information related to detection is discarded, which could lead to biased abundance estimates under certain conditions.
-Most of the code used to conduct analyses is shown in body of this manuscript.
-However, some code related to generating tables and figures is not shown, for brevity, although it can be accessed via https://github.com/tmeeha/inlaNMix.
+However, in the process of averaging, important information related to detection is discarded, which could lead to biased detection and abundance estimates under certain conditions.
+Much of the code used to conduct analyses is shown in the body of this manuscript.
+However, some repeated code, and code related to generating tables and figures, is not shown for brevity.
+All code can be accessed via https://github.com/tmeeha/inlaNMix.
 
@@
 
-The data simulated for this analysis was intended to represent a typical wildlife study.
+The data simulated for this analysis were intended to represent a typical wildlife study.
 To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites.
@@
 
-In this scenario, the abundance of the species is thought to vary with two site-level covariates (covariates 1 and 2) that represents habitat characteristics at a site and do not change appreciably over time.
-The detection probability is also believed to vary according to two covariates (covariates 1 and 3).
-The first covariate for detection, covariate 1, is the same site-level covariate 1 that affects abundance, although it has the opposite effect on detection.
-The other detection covariate, covariate 3, is a site-survey-year variable that could be related to weather conditions during a given count.
-
-Finally, as is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables.
-Overdispersion is generated using a negative binomial distribution for the count component of the model.
-The specific model parameters and their simulated values are: p.b0 = 1.0 (detection intercept), p.b1 = -2.0 (effect of detection covariate 1), p.b3 = 1.0 (effect of detection covariate 3), lam.b0 = 2.0 (abundance intercept), lam.b1 = 2.0 (effect of abundance covariate 1), lam.b2 = -3.0 (effect of abundance covariate 2), lam.b4 = 1.0 (effect of year on abundance), disp.size = 3.0 (size of the overdispersion parameter).
-All independent variables in the simulation are centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.
+In this scenario, the abundance of the species is thought to vary with two site-level covariates ($\lambda$ covariates 1 and 2) that represents habitat characteristics at a site and do not change appreciably over time.
+The detection probability is also believed to vary according to two covariates (*p* covariates 1 and 3).
+The first covariate for detection, *p* covariate 1, is the same site-level covariate 1 that affects abundance, although it has the opposite effect on detection.
+The other detection covariate, *p* covariate 3, is a site-survey-year variable that could be related to weather conditions during a given count.
+
+As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables.
+Overdispersion is generated and modeled using a negative binomial distribution for the count component of the model.
+The specific model parameters and their simulated values are: p.b0 = 1.0 (*p* intercept), p.b1 = -2.0 (effect of *p* covariate 1), p.b3 = 1.0 (effect of *p* covariate 3), lam.b0 = 2.0 ($\lambda$ intercept), lam.b1 = 2.0 (effect of $\lambda$ covariate 1), lam.b2 = -3.0 (effect of $\lambda$ covariate 2), lam.b4 = 1.0 (effect of year on $\lambda$), disp.size = 3.0 (size of the overdispersion parameter).
+All independent variables in the simulation vary are centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.
 
 We simulated data for this study using the following function.
-Note that the function produces two versions of one detection covariate, x.p.3 and x.p.3.arr, and two versions of the count matrix, y1 and y2.
+Note that the function produces two versions of detection covariate 3, x.p.3 and x.p.3.arr, and two versions of the count matrix, y1 and y2.
 x.p.3 is the site-survey-year variable described above.
-It is used to generate y2, which is used in excersize 2, below.
+It is used to generate y2, which is used in Example II.
 x.p.3.mean is derived from x.p.3, where values are unique to site and year, but averaged over surveys.
-It is used to generate y1, which is used here, in exercize 1.
+It is used to generate y1, which is used in Example I.
 
@@
 
-In addition to simulated data, we also demonstrate the use of R-INLA using a real dataset in exercise 3.
-This dataset comes from a study by Kery et al.
-(2005) and is publically available as part of the unmarked package.
-The dataset includes mallard duck counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors the abundance of breeding birds.
-In addition to counts, the dataset also includes 2 site-by-survey covariates related to detection (effort and survey date), and 3 site-level covariates related to abundance (route length, elevation, and forest cover).
-Full dataset details are given in Kery et al.
-(2002).
-
-##2. Exercise 1
+In addition to simulated data, we also demonstrate the use of R-INLA using a real dataset in Example III.
+This dataset comes from a study by K&#233;ry et al.
+(2005) and is publicly available as part of the unmarked package.
+The dataset includes mallard duck (*Anas platyrhynchos*) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance.
+In addition to counts, the dataset also includes 2 site-survey covariates related to detection (survey effort and survey date), and 3 site level covariates related to abundance (route length, elevation, and forest cover).
+Full dataset details are given in K&#233;ry et al.
+(2005).
+
+#2. Example I
 
 ###2.1 Goals
-In Excersize 1, we demonstrate the use of R-INLA and compare use and performance to a simular analyses using JAGS and unmarked.
-In this excercise, the forms of JAGS, unmarked, and INLA models match the data generating process.
-Specifically, we used x.p.3.mean to generate y1 and analyzed the data using models that use x.p.3.mean as a covariate.
-This exercise demonstrates the differences in package use, computation time, and estimation results.
-
-###2.2 JAGS model specification
+
+In Example I, we demonstrate the use of R-INLA and compare use and performance to similar analyses using JAGS and unmarked.
+In this exercise, the forms of JAGS, unmarked, and R-INLA models match the data generating process.
+Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data using models that use x.p.3.mean as a covariate.
+This exercise is intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three packages when the specified model is reasonably close to the data generating process.
+
+###2.2 Analysis with JAGS
 
 We first analyze the simulated data using JAGS.
-In defining the model, we specify a negative binomial model for the abundance component, and use vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and p.
+In defining the model, we specify a negative binomial model for the abundance component, and use vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and *p*.
 
@@
 ```
-
-###2.3 JAGS parameters, dataset, and initial values
 
@@
 
-###2.4 Run JAGS model via runjags and view results
-
 Finally, we set the run parameters and start the MCMC process.
@@
 Note that the recommended number of effective samples for particularly robust inference is closer to 6000 (Gong and Flegal 2016).
-So MCMC processing times reported here could be considered shorter than necessary.
+So MCMC processing times reported here could be considered low estimates.
 
@@
 
-```{r print jags out 1}
+```{r print jags out 1, eval=F, echo=F}
 # view summary
@@
 
-Mean parameter estimates from the JAGS model are reasonably close to, and not significantly different from, the input values used to generate the data (Table 1).
-The full marginal posterior distributions for model parameters are shown in Figure 1.
-The potential scale reduction factor for all variables was < 1.05, and the effective sample size for all variables was > 1640.
-The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 20 minutes.
-
-###2.5 Prepare data for unmarked
-
-Next, we prepare data for the unmarked package, which involved slight modification of the data in the sim.data object.
+Mean parameter estimates from the JAGS model are reasonably close to, and not significantly different from, the input values used to generate the data (Fig.
+1).
+The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2117.
+The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2162 seconds.
+
+###2.3 Analysis with unmarked
+
+Next, we prepare the simulated data for the unmarked analysis, which involves slight modification of the data in the sim.data object.
 Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by *J* column matrix.
-Each static, site-level variable was duplicated and stacked *K* times to form a single column vector.
+Each static, site level variable was duplicated and stacked *K* times to form a single-column vector.
 A column vector was created to identify each year in the stacked data.
-The site-by-year variable, x.p.3, was transformed from 2-dimensional matrix to a single column vector.
+The site-year variable, x.p.3.mean, was transformed from 2-dimensional matrix to a single column vector.
 Reformatted variables were then assembled in an unmarked data structure called an unmarked frame.
@@
 
-###2.6 Analyze model using unmarked and view results
-
 The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance.
-This was followed by an argument identifying the appropriate unmarked frame and the form of the count portion of the mixture model, negative binomial.
+This was followed by an argument identifying the appropriate unmarked frame, and the form of the count portion of the mixture model, negative binomial.
 
@@
 
-```{r print unmk out 1}
+```{r print unmk out 1, echo=F, eval=F}
 # view summary
@@
 
-Maximum liklihood estimates from the unmarked analysis are are also close to, and not significantly different from, input values (Table 1)  The analysis took approximately 120 seconds.
-
-###2.7 Prepare data for R-INLA
-
-Next, we prepare data for R-INLA, which involved slight modification of the data in the sim.data object, which produced data formatted for JAGS.
-The object counts.and.count.covs is an R-INLA object that includes an $I*K$ row by j column matrix of counts.
-Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for lambda.
-Finally, there are two $I*K$ length vector of values for the two static, site covariates of abundance, where the vector of values for i sites is stacked k times.
-In addition to the counts.and.count.covs object, we also define x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.
+Maximum likelihood estimates from the unmarked analysis are also close to, and not significantly different from, input values (Fig.
+1)  The unmarked estimates were produced in approximately 72 seconds.
+
+###2.4 Analysis with R-INLA
+
+Next, we prepare data for R-INLA, which uses data formats similar to those used by unmarked.
+The object counts.and.count.covs is an R-INLA object that includes an $I*K$ row by *J* column matrix of counts.
+Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for $\lambda$.
+Finally, there are two $I*K$ length vector of values for the two static, site covariates of abundance, where the vector of values for *I* sites is stacked *K* times.
+In addition to the counts.and.count.covs object, we also define x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.mean.
 
@@
 
-###2.8 Analyze model using R-INLA and view results
-
 The call to the inla() function includes several components.
@@
 On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.
-On the right side of the formula is a 1 to specify a global intercept for p and the two covariates for p.
+On the right side of the formula is a 1 to specify a global intercept for *p* and the two fixed covariates for *p*.
+Note that a wide range of random effects for *p* could be added to the right side of the formula (Rue et al.
+2013).
 The second argument describes the data, provided here as a list that corresponds with the model formula.
-Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture for the count component of the model.
-The fourth and fifth arguments specify the priors for the two model components, which are similar to those in the JAGS model.
+Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture.
+The fourth and fifth arguments specify the priors for the two model components.
+In this case, the priors are specified similarly to those in the JAGS model, but a variety of other options are available.
+Several other characteristics of the analysis can be modified using the call to inla().
+See Rue et al.
+(2013) for details.
 
@@
 
-```{r print inla out 1}
+```{r print inla out 1, echo=F, eval=F}
 # view summary
@@
 
-Mean parameter estimates from the R-INLA model are also close to, and not significantly different from, input values (Table 1)  Full marginal posterior distributions from R-INLA are also shown in Figure 1.
-The analysis took approximately 6 seconds.
+Mean parameter estimates from the R-INLA model are also close to, and not significantly different from, input values (Fig.
+1)  Full marginal posterior distributions from R-INLA are also shown in Fig.
+1.
+The analysis took approximately 7 seconds.
 
@@
 
-```{r print table, echo=F}
+```{r print table, echo=F, eval=F}
 print(kable(all.out.tab, align="c", caption="Parameter estimates from JAGS, unmarked, and R-INLA"))
@@
 
-###2.9 Notes on Exercise 1
-
-Exercise 1 shows several things.
-
-In demonstrating the use of R-INLA, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.
-Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for lambda are specified through an R-INLA object included on the left side of the formula.
-
-Regarding performance, both R-INLA and JAGS successfully extracted simulation input values.
-Figure 1 depicts marginal posterior distributions produced by JAGS and R-INLA.
-These posteriors derive from data resulting from one random manifestation of the input values.
-Thus we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.
-However, we do expect the input values to fall somewhere within the posterior distributions, which is what occurs here.
-Figure 1 shows that, for similarly specified models, R-INLA (dashed black lines) and JAGS (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.
-Where R-INLA and JAGS differed was in computing time.
-In this example, MCMC took 1,215 seconds to produce posteriors while INLA took 6 seconds – a 200-fold difference.
-This was the case despite the fact that the JAGS model was run in parallel with each of three MCMC chains simulated on a separate virtual core.
-If parallel computing had not been used, processing the JAGS model would have taken approximately twice as long.
-If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have been tripled again.
-
-```{r make plot 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1. Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines)."}
+```{r make plot 1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Marginal posteriors from JAGS (solid gray lines) and R-INLA (dashed black lines), along with maximum liklihood estimates (black circles) and 95% confidence intervals (horizontal black lines) from unmarked.
+True input values are represented by vertical black lines."}
 # get output from jags
@@
 d2$x <- plogis(d2$x)
-p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)
+col1 <- 1
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- plogis(c(d3,d4,d5))
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p5 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("p")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=plogis(1), col="gray20", lty=1)+
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.1.p
@@
 d2 <- as.data.frame(density(jags.df$bCov1P)[c(1,2)])
-p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)
+col1 <- 2
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- c(d3,d4,d5)
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p6 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 1")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=-2, col="gray20", lty=1)+
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.3.p
@@
 d2 <- as.data.frame(density(jags.df$bCov3P)[c(1,2)])
-p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)
+col1 <- 3
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- c(d3,d4,d5)
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p7 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Detection covariate 3")  + ylab(NULL) + theme_acbs() + geom_vline(xintercept=1, col="gray20", lty=1)+
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for overdisp
@@
 d2 <- as.data.frame(density(jags.df$overDisEst)[c(1,2)])
-p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)
+col1 <- 8
+d3 <- all.out.tab[col1, 5]
+d4 <- all.out.tab[col1, 4]
+d5 <- all.out.tab[col1, 6]
+d6 <- exp(c(d3,d4,d5))
+d7 <- data.frame(y=c(0,0,0), x=d6)
+p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)+
+geom_line(data=d7, aes(y=y, x=x)) +
+geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # plot all
@@
 
-##3. Example 2, Simulated Data
+###2.5 Example I summary
+
+Example I demonstrated basic use of R-INLA and highlighted similarities and differences between it and two other commonly used approaches.
+In demonstrating the use of R-INLA, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.
+Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for $\lambda$ are specified through an R-INLA object included on the left side of the formula, and covariates for *p* are specified on the right side of the formula.
+Data format and model specification syntax of R-INLA are not too different from unmarked, which are both considerably different from JAGS and other BUGS-oriented MCMC software.
+
+Regarding performance, R-INLA, unmarked, and JAGS successfully extracted simulation input values.
+Fig.
+1 depicts marginal posterior distributions produced by JAGS and R-INLA, and 95% confidence intervals from unmarked.
+These results derive from data resulting from one random manifestation of the input values.
+Thus, we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.
+However, we do expect the input values to fall somewhere within the posterior distributions and 95% confidence intervals, which is what occurs here.
+Figure 1 shows that, for similarly specified models, R-INLA (dashed black lines) and JAGS (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.
+
+Where R-INLA, unmarked, and JAGS differed was in computing time.
+In this example, R-INLA took approximately 7 seconds, unmarked took approximately 72 seconds, and JAGS took 2,162 seconds to produce results.
+Thus, R-INLA was approximately 10 times faster than unmarked and 200 times faster than JAGS.
+This was the case despite the fact that unmarked uses ML algorithms and the JAGS model was run in parallel with each of three MCMC chains simulated on a separate virtual core.
+If parallel computing had not been used with JAGS, processing the JAGS model would have taken approximately twice as long.
+If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have been tripled again.
+
+In sum, when compared to other tools, R-INLA is relatively easy to implement and produces accurate estimates of Bayesian posteriors relatively quickly.
+This statement assumes that the data generating process can be captured accurately in model specification.
+However, as mentioned above, certain models can not be specified using R-INLA.
+For the data in Example I the count matrix was produced using a detection covariate that was averaged to the site-year level.
+This averaged covariate was subsequently used in the analysis.
+But what happens when the site-survey-year covariate is an important component of the data generating process, and it can't be specified in the model in this form?
+This is the question explored in Example II.
+
+#3. Example II
 
 ###3.1 Goals
-In Example 2, we show the consequences of not being able to specify a survey level covariate.
-
-There are also other differences between the two approaches that, in some cases, will favor the use of JAGS and other BUGS oriented approaches, despite their long processing times.
-BUGS based approaches allow users ultimate flexibility in specifying models, whereas, when using R-INLA, only a subset of possible models can be specified.
-For example, when using R-INLA, random effects cannot be specified for $\lambda$.
-However, it is possible to specify a wide variety of random effects for p, including simple exchangeable intercepts as well as spatially- and temporally-structured random effects.
-As another example, when using R-INLA, covariates for $\lambda$ and p need to have the same degree of granularity, e.g., it is not possible to have site-year covariates for lambda and site-survey-year covariates for p.
-This limitation might become important if survey level covariates are not somewhat controlled by the sampling design.
-In this case, using the means of survey-level covariates per site and year will, at least, allows users to correct for any systematic differences in site-survey-year covariates across sites that might confound estimates of other site-level effects.
+
+In Example 2, we show the consequences of not being able to specify a site-survey-year covariate under a range of conditions.
+Specifically, we conducted a Monte Carlo simulation where, for each iteration, the count matrix for the analysis, y2, was generated with the data generating function described in 1.3, using the site-survey-year covariate x.p.3.
+The count data were then analyzed with two JAGS models.
+One model incorporated an averaged site-survey x.p.3.mean as a covariate, such as in 2.1.
+The other model incorporated the site-survey-year covariate as a fixed effect instead.
+In each iteration we varied the size of the effect for x.p.3.
+We suspected that the simpler model, with x.p.3.mean, would yield biased estimates when the effect of x.p.3 was large and unbiased estimates when the effect was small.
+
+###3.2 Analysis with JAGS
+
+Given the long processing time associated with JAGS models, we only ran 3000 iterations (after 1000 adaptive iterations and 1000 burn-in iterations) during each simulation.
+This number is not sufficient for making inference on marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means.
+In each of the 50 iterations the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3.
+The results of the simulations are depicted in Fig.
+2.
 
@@
 
-```{r explore sim results, echo=FALSE}
+```{r explore sim results, fig.pos="h", echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.
+Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.
+Parameter name and true input value is given in the strip across the top of each panel."}
 # explore results
@@
 to.string <- as_labeller(par_names)
-ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=0.5) +
+ggplot(data=sim.out, aes(x=p.b3, y=diffs, colour=simtype, linetype=simtype)) + geom_point(pch=1, size=1.5) +
   facet_wrap(~factor(par), scales="fixed",labeller=to.string) +
@@
 
-###3.2  Results
-
-This is what happened in this section.
-
-##4. Example 3, Real Data
+###3.3  Example II summary
+
+As expected, biases in estimates of abundance ($\lambda$ intercept) and the effects of abundance covariates (specifically $\lambda$ covariate 2) increased with the effect strength of the site-survey-year covariate for *p*.
+When the effect size was small, with an absolute value less than 1, the bias was minimal.
+When the effect size was large, with an absolute value greater than 2, the bias was large.
+When interpreting the effect size, bear in mind that x.p.3 ranged from -0.5 to 0.5.
+
+#4. Example III
 
@@
 
-We wanted to show real data.
-
-###4.1 unmarked
-
-```{r second unmarked analysis}
+In Example III, we explored the performance of R-INLA using real data, a publicly available dataset on mallard duck abundance in Switzerland in 2002.
+By incorporating real data, we hoped to evaluate the performance of R-INLA using data that were not predictable by design and the practical impacts of not being able to specify survey level covariates in R-INLA.
+The dataset is available as a demonstration dataset in unmarked, so we compared the performance of R-INLA and unmarked in Example III.
+
+###4.1 Analysis with unmarked
+
+The unmarked model was specified with linear effects of transect length, elevation, and forest cover on abundance, and a linear effect and quadratic effect of sampling intensity and sampling date, respectively, on detection.
+Both sampling intensity and sampling date were site-survey covariates.
+
+```{r second unmarked analysis, message=FALSE, warning=FALSE}
 # get real mallard data from unmarked package
@@
 
-```{r summary for unmk out 2}
+```{r summary for unmk out 2, echo=F, eval=F}
 summary(mallard.out.unmk)
@@
 
-###4.2 INLA
-
-```{r second inla analysis}
+###4.2 Analysis with INLA
+
+The R-INLA model used a similar model structure.
+The only difference was that the R-INLA model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead.
+
+```{r second inla analysis, message=FALSE, warning=FALSE}
 # format inla data
@@
 mallard.forest.inla <- mallard.site[,3]
-mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T)
+mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # average per site
 mallard.ivel.inla[is.na(mallard.ivel.inla)] <- mean(mallard.ivel.inla, na.rm=T)
-mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T)
+mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # average per site
 mallard.date2.inla <- mallard.date.inla^2
@@
 
-```{r summary for out inla 2}
+```{r summary for out inla 2, echo=F, eval=F}
 summary(mallard.out.inla)
@@
 
-```{r exercize three plot, echo=F}
+```{r exercize three plot, fig.pos="h", echo=F, fig.cap="Estimates and 95% confidence intervals from unmarked (gray circles and lines) and posterior means and 95% credible intervals from R-INLA (black lines) from an N-mixture model of mallard duck abundance.
+The unmarked model specified unique site-survey covariates for survey effort and survey date, while the R-INLA model specified site-average versions.
+A value of zero is depicted by the vertical gray line."}
 # get unmk out
@@
 colnames(all.out.tab.2) <- c("Estimate","Lower", "Upper")
-all.out.tab.2$Parameter <- factor(rep(c("P Int", "P Ivel", "P Date", "P Date Sq",
-                           "Lambda Int", "Lambda Length", "Lambda Elev", "Lambda Forest", 
+all.out.tab.2$Parameter <- factor(rep(c("Detection intercept", "Survey intensity", 
+                                        "Survey date", "S. date squared", "Abundance intercept",
+                                        "Transect length", "Transect elevation", "Forest cover",
                                         "Overdispersion"), 2))
@@
   geom_errorbar(aes(ymin = Lower, ymax = Upper), position=pd) + 
-  theme_acbs() + scale_color_manual(values=c("black","gray50")) + coord_flip()
-```
-
-This is what we found.
-
-##5. Discussion
-
-The purpose of this study was to (1) demonstrate the use of the R-INLA package to analyze N-mixture models used by wildlife biologists and (2) compare performance of R-INLA to another common approach, JAGS via the runjags package, which uses MCMC methods.
-
-##6.0 Appendix A
-
-###6.1 Likelihood
-
-Would you like to add stuff here about the likelihood functions and the iterative procedure for solving them?
-
-##7.0 References
+  theme_acbs() + scale_color_manual(values=c("black","gray50")) + coord_flip() +
+  geom_hline(yintercept=0, col="gray30") + 
+  xlab("Estimated value (95% credible or confidence interval)") +
+  ylab("Model parameter")
+```
+
+The 95% credible intervals from the R-INLA model overlapped considerably with the 95% confidence intervals from unmarked model, so parameter estimates were not significantly different from one another.
+Also, the parameters with values significantly different from zero were the same regardless of technique.
+Using both techniques, detection decreased linearly as the season progressed and abundance decreased as linear functions of increased forest cover and elevation gain.
+Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for unmarked and site-averaged detection covariates were used for R-INLA.
+Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, might lead to similar results regardless of which covariate is used.
+Conclusions may have been different with a different real dataset where detection covariates had very strong effects.
+
+#5. Discussion
+
+The purpose of this study was to demonstrate the use of the R-INLA package to analyze N-mixture models and to compare performance of R-INLA to two other common approaches, JAGS (via the runjags package), which uses MCMC methods and allows Bayesian inference (Lunn et al.
+2013), and unmarked, which uses maximum likelihood and allows frequentist inference (Fiske et al.
+2011).
+
+Results demonstrate that R-INLA can be a complementary tool in the wildlife biologist's analytical tool kit.
+Strengths of R-INLA include Bayesian inference, based on highly accurate approximations of posterior distributions, which are derived 200 times faster than MCMC methods, where models are specified using a syntax that should be familiar to R users, and where data are formatted in a straightforward way with relatively few lines of code.
+The straightforward model syntax and data format could help remove  barriers to adoption of N-mixture models for biologists who are not committed to learning the BUGS syntax.
+The substantial decrease in computation time should allow a wider variety of model and variable selection techniques (e.g., cross validation and model averageing), ones that are not commonly used in an MCMC context due to practical issues related to computing time (K&#233;ry and Schaub 2011).
+
+Limitations of R-INLA are mainly related to the more restricted set of N-mixture models that can be specified.
+Of the approaches explored here, BUGS-based approaches allow users ultimate flexibility in specifying models.
+For example, with JAGS, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available (Joseph et al.
+2009, Martin et al.
+2011), and a variety of random effects can be specified for both $\lambda$ and *p* (K&#233;ry and Schaub 2011).
+In comparison, the current version of R-INLA does not handle survey-level covariates, employs only Poisson-binomial and negative binomial-binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for *p* only.
+In cases where survey-level covariates are particularly strong and not controlled in the sampling design, R-INLA will not be the appropriate tool (Fig.
+2).
+
+When compared to unmarked, the R-INLA approach is similar in regards to model syntax and data format.
+The approaches are also similar in that both yield results much faster than MCMC.
+The two approaches differ in that R-INLA is roughly 10 times faster than unmarked, possibly due to the different approach used to compute model likelihoods (see Appendix).
+They also differ in that unmarked can accomodate survey level covariates and can analyze dynamic N-mixture models (Dail and Madsen 2011).
+
+In conclusion, JAGS (and WinBUGS, OpenBUGS), unmarked, and R-INLA all allow users to employ N-mixture models for estimating wildlife abundance while accounting for imperfect detection.
+Each method has its strengths and limitations.
+R-INLA appears to be an attractive option when survey level covariates are not essential, familiar model syntax and data format are desired, fast computing time is needed, and Bayesian inference is preferred.
+
+#6. Appendix
+
+###Recursive computations of the likelihood for N-mixture models
+
+The likelihood for the simplest case is
+\begin{displaymath}
+    \text{Prob}(y) = \sum_{n = y}^{\infty}
+    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
+\end{displaymath}
+where $\text{Poisson}(n; \lambda)$ is the density for the Poisson
+distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and
+$\text{binomial}(y; n, p)$ is the density for the binomial distribution
+with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
+Although the likelihood can be computed directly when replacing the
+infinite limit with a finite value, we will demonstrate here that we
+can easily evaluate it using a recursive algorithm that is both faster
+and more numerical stable. The same idea is also applicable to the
+negative binomial case, and the case where we have replicated
+observations of the same ``$n$''. We leave it to the reader to derive
+these straight forward extensions.
+
+The key observation is that both the Poisson and the binomial
+distribution can be evaluated recursively in $n$,
+\begin{displaymath}
+    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
+\end{displaymath}
+
+and
+
+\begin{displaymath}
+    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
+\end{displaymath}
+
+and then also for the Poisson-binomial product
+
+\begin{displaymath}
+    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
+    =
+    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
+    \frac{\lambda}{n-y}(1-p).
+\end{displaymath}
+
+If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make
+use of this recursive form to express the likelihood with a finite
+upper limit as
+
+\begin{eqnarray}
+    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
+                       \text{Poisson}(n ; \lambda)\;
+                      \text{binomial}(y;  n, p) \nonumber\\
+                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
+                       \Big\{ 1 + f_1 + f_1f_2 + 
+                    \ldots
+                      +f_1\cdots f_{n_\text{max}}
+                       \Big\} \nonumber\\
+                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
+                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
+\end{eqnarray}
+
+The log-likelihood can then evaluated using the following simple \texttt{R}-code
+
+```{r, eval=F}
+fac = 1; ff = lambda * (1-p)
+for (i in (n.max - y):1) fac = 1 + fac * ff / i
+log.L = dpois(y, lambda, log=TRUE) +
+        dbinom(y, y, p, log=TRUE) + log(fac)
+```
+
+Since this evaluation is recursive in decreasing $n$, we have to chose
+the upper limit $n_\text{max}$ up-front, for example as an integer
+larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is
+small. Note that we are computing \texttt{fac} starting with the
+smallest contributions, which are more numerical stable.
+
+#7. References
 
@@
 
-Kery, M. 2010. Introduction to WinBUGS for ecologists: Bayesian approach to regression, ANOVA, mixed models and related analyses. Academic Press.
-
-Kery, M., R. M. Dorazio, L. Soldaat, A. Van Strien, A. Zuiderwijk, and J. A. Royle. 2009. Trend estimation in populations with imperfect detection. Journal of Applied Ecology 46:1163–1172.
-
-Kery, M., and J. A. Royle. 2010. Hierarchical modelling and estimation of abundance and population trends in metapopulation designs. Journal of Animal Ecology 79:453–461.
-
-Kery, M., and J. A. Royle. 2015. Applied hierarchical modeling in ecology: analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models. Academic Press.
-
-Kery, M., J. A. Royle, and H. Schmid. 2005. Modeling avian abundance from replicated counts using binomial mixture models. Ecological applications 15:1450–1461.
-
-Kery, M., and M. Schaub. 2011. Bayesian population analysis using WinBUGS: a hierarchical perspective. Academic Press.
+K&#233;ry, M. 2010. Introduction to WinBUGS for ecologists: Bayesian approach to regression, ANOVA, mixed models and related analyses. Academic Press.
+
+K&#233;ry, M., R. M. Dorazio, L. Soldaat, A. Van Strien, A. Zuiderwijk, and J. A. Royle. 2009. Trend estimation in populations with imperfect detection. Journal of Applied Ecology 46:1163–1172.
+
+K&#233;ry, M., and J. A. Royle. 2010. Hierarchical modelling and estimation of abundance and population trends in metapopulation designs. Journal of Animal Ecology 79:453–461.
+
+K&#233;ry, M., and J.
+A.
+Royle.
+2015.
+Applied hierarchical modeling in ecology: analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models.
+Academic Press.
+
+K&#233;ry, M., J. A. Royle, and H. Schmid. 2005. Modeling avian abundance from replicated counts using binomial mixture models. Ecological applications 15:1450–1461.
+
+K&#233;ry, M., and M. Schaub. 2011. Bayesian population analysis using WinBUGS: a hierarchical perspective. Academic Press.
 
@@
 Edwards, M.
-Kéry, and B.
+K&#233;ry, and B.
 Gardner.

### 2017-04-18 rmark_doc_inla_nmixnb_8.Rmd
Incorporated Nicole's comments into manuscript

@@
 
-Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture or binomial mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, K&#233;ry et al.
+Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture model (Royle and Nichols 2003, Dodd and Dorazio 2004, Royle 2004, K&#233;ry et al.
 2005).
-A simple form of an N-mixture model describes observed counts *Y* at site *i* during survey *j* as coming from a binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$.
+A simple form of an N-mixture model, a binomial mixture model, describes observed counts *Y* at site *i* during survey *j* as coming from a binomial distribution with parameters for abundance *N* and detection probability *p*, where *N* per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$.
 Specifically,
@@
 However, in the process of averaging, important information related to detection is discarded, which could lead to biased detection and abundance estimates under certain conditions.
+
 Much of the code used to conduct analyses is shown in the body of this manuscript.
@@
 (2005) and is publicly available as part of the unmarked package.
-The dataset includes mallard duck (*Anas platyrhynchos*) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance.
+The dataset includes mallard duck (*Anas platyrhynchos*) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance (Monitoring H&#228;ufige Brutv&#246;gel).
 In addition to counts, the dataset also includes 2 site-survey covariates related to detection (survey effort and survey date), and 3 site level covariates related to abundance (route length, elevation, and forest cover).
@@
 d7 <- data.frame(y=c(0,0,0), x=d6)
-p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdisperison parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)+
+p8 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) + geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("Overdispersion parameter") + ylab(NULL) + theme_acbs() + geom_vline(xintercept=3, col="gray20", lty=1)+
 geom_line(data=d7, aes(y=y, x=x)) +
@@
     \text{Prob}(y) = \sum_{n = y}^{\infty}
-    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
+    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p),
 \end{displaymath}
@@
 \begin{displaymath}
-    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
+    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p).
 \end{displaymath}
 
-and then also for the Poisson-binomial product
+And then also for the Poisson-binomial product,
 
@@
                    &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
-                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
+                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}.\nonumber
 \end{eqnarray}

### 2017-04-20 meehan_et_al_inla_nmix_jss/article.tex
Moved old stuff to archive and added new stuff
I moved all the old material to an archive folder.  I also added a new
top level folder for a JSS manuscript.  There is a TEX file in there
(article.tex) that is the current working document.  There is also a PDF
copy (article.pdf), and other latex building block stuff as well. @hrue
, @nicolelmichel

%% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[article]{jss}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National Audubon Society \And
        Nicole L. Michel\\National Audubon Society \And
        H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

%% for pretty printing
%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, H\"{a}vard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

%% an abstract and keywords
\Abstract{
  Successful management of wildlife populations requires accurate estimates of abundance.  Abundance estimates can be confounded by imperfect detection during wildlife surveys. Development of N-mixture models has facilitated quantification of detectability and has improved our ability to produce accurate abundance estimates. The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo methods and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference. Our results suggest that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) survey level covariates are not essential, (2) familiar model syntax and data format are desired (relative to other R packages), (3) fast computing time is necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.

}
\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \proglang{R}, \proglang{BUGS}}
%% without formatting
\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}

%% publication information
%% NOTE: This can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  H\"{a}vard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, SAU\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% start main document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%% introduction
\section[Introduction]{Introduction}
\subsection[Background]{Background}
Successful management of wildlife species requires accurate estimates of abundance \citep{Yoccoz_Nichols_Boulinier_2001}.  One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}.  Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure.  The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}.  One such sampling scheme, termed a "metapopulation design" \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.  If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.  This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Denes_Silveira_Beissinger_2015}.

Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture model \citep{Royle_Nichols_2003, Dodd_Dorazio_2004, Royle_2004, Kery_Royle_Schmid_2005}.  A simple form of an N-mixture model, a binomial mixture model, describes observed counts $Y$ at site $i$ during survey $j$ as coming from a binomial distribution with parameters for abundance $N$ and detection probability $p$, where $N$ per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$. Specifically,

$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim binomial(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, $p$ is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.  In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim binomial(N_{i,k}, p_{i,k})$.  Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.  There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.  Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.  We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for a nice overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology.  However, there are practical issues that sometimes act as barriers to adoption.  Much of the development of N-mixture models, and many of the examples in the wildlife literature, implement models using Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.  These are extremely powerful and flexible platforms for analyzing Bayesian models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax.  While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment.  Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.

There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.  The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a Frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.  The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} syntax.  Models are analyzed using a Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach.  The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.  However, it comes at a cost -- the loss of the intuitive inferential framework associated with Bayesian analysis.

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} for \proglang{R}.  The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC.  Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in R.  Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low.  The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

\subsection[Overall objectives]{Overall objectives}
The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \pkg{R-INLA} package. In the process, we employ simulated and real count datasets, and analyze them using \proglang{JAGS}, via the \pkg{runjags} package \citep{Denwood_2016} for \proglang{R}, the \pkg{unmarked} package, and the \pkg{R-INLA} package. In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance. We also explore a limitation of the \pkg{R-INLA} approach related to model specification.  Specifically, while it is possible to specify survey level covariates for detection using \proglang{JAGS} and \pkg{unmarked}, this is not possible using \pkg{R-INLA}.  Rather, survey level covariates of detection must be averaged to the site or site-year level.  Using an averaged detection covariate does allow accounting for site-level differences in survey conditions should they occur.  However, in the process of averaging, important information related to detection is discarded, which could lead to biased detection and abundance estimates under certain conditions.

Much of the code used to conduct analyses is shown in the body of this manuscript. However, some repeated code, and code related to generating tables and figures, is not shown for brevity.  All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}.

\subsection[Simulated data]{Simulated data}
The data simulated for this analysis were intended to represent a typical wildlife study. To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites. At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season when birds are likely to be singing. In order to estimate a trend in abundance over time, these repeated surveys are conducted over a 9-year period.

In this scenario, the abundance of the species is thought to vary with two site-level covariates ($\lambda$ covariates 1 and 2) that represent habitat characteristics at a site and do not change appreciably over time. The detection probability is also believed to vary according to two covariates ($p$ covariates 1 and 3). The first covariate for detection, $p$ covariate 1, is the same site-level covariate 1 that affects abundance, although it has the opposite effect on detection. The other detection covariate, $p$ covariate 3, is a site-survey-year variable that could be related to weather conditions during an individual survey.

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 ($p$ intercept), p.b1 = -2.0 (effect of $p$ covariate 1), p.b3 = 1.0 (effect of $p$ covariate 3), lam.b0 = 2.0 ($\lambda$ intercept), lam.b1 = 2.0 (effect of $\lambda$ covariate 1), lam.b2 = -3.0 (effect of $\lambda$ covariate 2), lam.b4 = 1.0 (effect of year on $\lambda$), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

We simulated data for this study using the following function.  Note that the function produces two versions of detection covariate 3, x.p.3 and x.p.3.mean, and two versions of the count matrix, y1 and y2.  x.p.3 is the site-survey-year variable described above.  It is used to generate y2, which is used in Example II.  x.p.3.mean is derived from x.p.3, where values are unique to site and year, but averaged over surveys.  It is used to generate y1, which is used in Example I.

\begin{Code}
data4sim <- function(n.sites = 72,   # number study sites
                     n.surveys = 3,  # short term replicates
                     n.years = 9,    # number years
                     lam.b0 = 2.0,   # intercept for log lambda
                     lam.b1 = 2.0,   # slope for log lambda, covariate 1
                     lam.b2 = -3.0,  # slope for log lambda, covariate 2
                     lam.b4 = 1.0,   # slope for log lambda, year
                     p.b0 = 1.0,     # intercept for logit p
                     p.b1 = -2.0,    # slope for logit p, covariate 1
                     p.b3 = 1.0,     # slope for logit p covariate 3
                     disp.size = 3.0 # size of the overdisperison
                     ){
  # setup
  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
  N.tr <- array(dim = c(n.sites, n.years))           # array for true abund
  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex1 counts
  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex2 counts
  # abundance covariate values
  x.lam.1 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years))
  x.lam.2 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
                   dim=c(n.sites, n.years))
  yrs <- 1:n.years; yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
  yr <- array(rep(yrs, each=n.sites), dim=c(n.sites, n.years))
  # fill abundance array
  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
  for(i in 1:n.sites){
    for(k in 1:n.years){
    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
  }}
  # detection covariate values
  x.p.1 <- array(x.lam.1[,1], dim=c(n.sites,n.surveys,n.years))
  x.p.3 <- array(as.numeric(scale(runif(n=n.sites*n.surveys*n.years, -0.5, 0.5),
                                  scale=F)), dim=c(n.sites,n.surveys,n.years))
  # average x.p.3 per site-year
  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
  out1 <- c()
  for(k in 1:n.years){
   chunk1 <- x.p.3.mean[,k]
   chunk2 <- rep(chunk1, n.surveys)
   out1 <- c(out1, chunk2)
  }
  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
  # fill count array with site-yr x.p.3
  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y1[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr1[i,j,k])
  }}}
  # fill count array with site-surv-yr x.p.3
  p.tr2 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3)
  for (i in 1:n.sites){
    for (k in 1:n.years){
      for (j in 1:n.surveys){
        y2[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr2[i,j,k])
  }}}
  # return data
  return(list(n.sites=n.sites, n.surveys=n.surveys, n.years=n.years,
              x.p.1=x.p.1[,1,1], x.p.3=x.p.3, x.p.3.mean=x.p.3.mean,
              x.p.3.arr=x.p.3.arr, x.lam.1=x.lam.1[,1],
              x.lam.2=x.lam.2[,1], yr=yr[1,], y1=y1, y2=y2,
              lam.tr=lam.tr, N.tr=N.tr))
} #end function

# create reproducible datasets for exercises 1 and 2
set.seed(12345)
sim.data <- data4sim()
\end{Code}

\subsection[Real data]{Real data}
In addition to simulated data, we also demonstrate the use of \pkg{R-INLA} with a real dataset in Example III.  This dataset comes from a study by \cite{Kery_Royle_Schmid_2005} and is publicly available as part of the \pkg{unmarked} package.  The dataset includes mallard duck (\emph{Anas platyrhynchos}) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance (Monitoring H\"{a}ufige Brutv\"{o}gel). In addition to counts, the dataset also includes 2 site-survey covariates related to detection (survey effort and survey date), and 3 site level covariates related to abundance (route length, elevation, and forest cover).  Full dataset details are given in \cite{Kery_Royle_Schmid_2005}.

%% example 1
\section[Example I]{Example I}
\subsection[Goals]{Goals}
In Example I, we demonstrate the use of \pkg{R-INLA} and compare use and performance to similar analyses using \proglang{JAGS} and \pkg{unmarked}.  In this exercise, the functional forms of \proglang{JAGS}, \pkg{unmarked}, and \pkg{R-INLA} models match the data generating process.  Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate.  This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three packages when the specified model was reasonably close to the data generating process.

\subsection[Analysis with JAGS]{Analysis with JAGS}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a negative binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

\begin{Code}
jags.model.string <- "
model {
  # priors
  intP ~ dnorm(0, 0.01)       # detection intercept
  bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
  bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
  intLam ~ dnorm(0, 0.01)     # lambda intercept
  bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
  bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
  bYr ~ dnorm(0, 0.01)        # year effect
  overDisEst ~ dunif(0, 5)    # overdispersion size
  # abundance component
  for (k in 1:nYears){
    for (i in 1:nSites){
      N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
      prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
      log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
                          (bYr * yr[k])
  # detection component
      for (j in 1:nSurveys){
        y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
        p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
        lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
      } # close j loop
    } # close i loop
  } # close k loop
} # close model loop
"
}
\end{Code}

Next, we defined the parameters to be monitored during the MCMC runs, bundled the data for \proglang{JAGS}, and created a function for drawing random initial values for the model parameters.  The initial values for abundance were made to avoid values of "NA" and zero, as these would cause computational problems.

\begin{Code}
# parameters to monitor
params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
            "bYr", "overDisEst")
# jags data
jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
             nYears = sim.data$n.years)
# initial values
N.init <- sim.data$y1 # initial count values
N.init[is.na(N.init)] <- 1 # clean up NA's
N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
                     overDisEst = runif(1, 0.5, 2.5))
\end{Code}

Finally, we set the run parameters and started the MCMC process. Run parameters were chosen such that MCMC diagnostics indicated converged chains (Gelman-Rubin statistics < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).  Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}.  So MCMC processing times reported here could be considered low estimates.

\begin{Code}
# set run parameters
nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10
# run jags
out.jags <- run.jags(model = jags.model.string, data = jags.data,
                     monitor = params, n.chains = nc, inits = inits,
                     burnin = nb, adapt = na, sample = ni, thin = nt,
                     modules = "glm on", method = "parallel")
summary(out.jags)
\end{Code}

Mean parameter estimates from the \proglang{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig. \ref{fig:fig1}).  The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2410.  The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.

%% figure 1
\begin{figure}
  \includegraphics[width=\linewidth]{fig1.jpeg}
  \caption{Marginal posteriors from \proglang{JAGS} (solid gray lines) and \pkg{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \pkg{unmarked}.  True input values are represented by vertical black lines.}
  \label{fig:fig1}
\end{figure}
%% end figure 1

\subsection[Analysis with unmarked]{Analysis with unmarked}
Next, we prepared the simulated data for the \pkg{unmarked} analysis, which involved slight modification of the data in the sim.data object.  Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by $J$ column matrix.  Each static site level variable was duplicated and stacked $K$ times to form a single-column vector. A column vector was created to identify each year in the stacked data.  The site-year variable, x.p.3.mean, was transformed from 2-dimensional matrix to a single column vector.  Reformatted variables were then assembled in an \pkg{unmarked} data structure called an unmarked frame.

\begin{Code}
# format count data
y.unmk <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.unmk <- rbind(y.unmk, y.chunk)
}
# format covariates
x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.unmk <- c(sim.data$x.p.3.mean)
site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
                             x.p.1.unmk, x.p.3.unmk)
# make unmarked pcount frame
unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
\end{Code}

The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance.  This was followed by an argument identifying the appropriate unmarked frame, and the form of the mixture model, negative binomial-binomial.

\begin{Code}
# run unmk model
out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
                   ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
                    data = unmk.data, mixture = "NB")
summary(out.unmk)
\end{Code}

Maximum likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1})  The \pkg{unmarked} estimates were produced in approximately 77 seconds.

\subsection[Analysis with R-INLA]{Analysis with R-INLA}
Next, we prepared data for \pkg{R-INLA}, which uses data formats similar to those used by \pkg{unmarked}. The object counts.and.count.covs is an \pkg{R-INLA} object that includes an $I*K$ row by $J$ column matrix of counts. Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for $\lambda$. Finally, there are two $I*K$ length vector of values for the two static site covariates of abundance, where the vector of values for $I$ sites is stacked $K$ times.  In addition to the counts.and.count.covs object, we also defined x.p.1.inla, which is a copy of x.lam.1.inla, and x.p.3.inla, which is the $I*K$ length vector corresponding with x.p.3.mean.

\begin{Code}
# format count data
y.inla <- sim.data$y1[ , , 1]
for(i in 2:sim.data$n.years){
  y.chunk <- sim.data$y1[ , , i]
  y.inla <- rbind(y.inla, y.chunk)
}
# format covariates
x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
x.p.3.inla <- c(sim.data$x.p.3.mean)
# make inla.mdata object
counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, x.lam.2.inla, yr.inla)
\end{Code}

The call to the inla() function includes several components.  First is the model statement.  On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for $p$ and the two fixed covariates for $p$.  Note that a wide range of random effects for $p$ could be added to the right side of the formula \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}.  The second argument describes the data, provided here as a list that corresponds with the model formula.  Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture.  The fourth and fifth arguments specify the priors for the two model components.  In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available. Several other characteristics of the analysis can be modified using the call to inla().  See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.

\begin{Code}
# run inla model
out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)),
                                          theta5 = list(prior = "flat",
                                                        param = numeric()))))
summary(out.inla)
\end{Code}

Mean parameter estimates from the \pkg{R-INLA} model are also close to, and not significantly different from, input values (Fig. \ref{fig:fig1})  Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig. \ref{fig:fig1}.  The analysis took approximately 6 seconds.

\subsection[Example I summary]{Example I summary}
Example I demonstrated basic use of \pkg{R-INLA} and highlighted similarities and differences between it and two other commonly used approaches.  In demonstrating the use of \pkg{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.  Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula. Data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.

Regarding performance, \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} successfully extracted simulation input values.  Fig. \ref{fig:fig1} depicts marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and 95\% confidence intervals from \pkg{unmarked}.   These results derive from data resulting from one random manifestation of the input values.  Thus, we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.  However, we do expect the input values to fall somewhere within the posterior distributions and 95\% confidence intervals, which is what occurs here.  Fig. \ref{fig:fig1} shows that, for similarly specified models, \pkg{R-INLA} (dashed black lines) and \proglang{JAGS} (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.

Where \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} differed was in computing time.  In this example, \pkg{R-INLA} took 6 seconds, \pkg{unmarked} took 77 seconds, and \proglang{JAGS} took 2104 seconds to produce results. Thus, \pkg{R-INLA} was approximately 10 times faster than \pkg{unmarked} and 300 times faster than \proglang{JAGS}.  This was the case despite the fact that \pkg{unmarked} uses ML algorithms and the \proglang{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core.  If parallel computing had not been used with \proglang{JAGS}, processing the \proglang{JAGS} model would have taken approximately twice as long.  If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have doubled again.

In sum, when compared to other tools, \pkg{R-INLA} is relatively easy to implement and produces accurate estimates of Bayesian posteriors very quickly. Its utility depends on the degree to which the data generating process can be captured accurately in model specification.  However, as mentioned above, certain models can not be specified using \pkg{R-INLA}.  For the data in Example I, the count matrix was produced using a detection covariate that was averaged to the site-year level.  This averaged covariate was subsequently used in the analysis. But what happens when the site-survey-year covariate is an important component of the data generating process, and it can't be specified in the model in this form?  This is the question explored in Example II.

%% example 2
\section[Example II]{Example II}
\subsection[Goals]{Goals}
In Example 2, we show the consequences of not being able to specify a site-survey-year covariate under a range of conditions.  Specifically, we conducted a Monte Carlo simulation where, for each iteration, the count matrix for the analysis, y2, was generated with the data generating function described in 1.3, using the site-survey-year covariate x.p.3.  The count data were then analyzed with two \proglang{JAGS} models.  One model incorporated an averaged site-survey x.p.3.mean as a covariate, such as in 2.1.  The other model incorporated the site-survey-year covariate instead.  In each iteration we randomly varied the size of the effect for x.p.3.  We suspected that the simpler model, with x.p.3.mean, would yield biased estimates when the effect of x.p.3 was relatively large, and unbiased estimates when the effect was relatively small.

\subsection[Analysis with JAGS]{Analysis with JAGS}
Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (after 1000 adaptive iterations and 1000 burn-in iterations) during each simulation.  This number is not sufficient for drawing inference from marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means.  In each of the 50 iterations the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3. The results of the simulations are depicted in Fig. \ref{fig:fig2}.

%% figure 2
\begin{figure}
  \includegraphics[width=\linewidth]{fig2.jpeg}
  \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.  Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.  Parameter name and true input value is given in the strip across the top of each panel.}
  \label{fig:fig2}
\end{figure}
%% end figure 2

\subsection[Example II summary]{Example II summary}
As expected, biases in estimates of abundance (abundance intercept) and the effects of abundance covariates (specifically, abundance covariate 1) increased with the effect strength of the site-survey-year covariate for $p$. When the effect size was small, with an absolute value less than 1, the bias was minimal.  When the effect size was large, with an absolute value greater than 2, the bias was large.  When interpreting the effect size, bear in mind that x.p.3 ranged from -0.5 to 0.5.

%% example 3
\section[Example III]{Example III}
\subsection[Goals]{Goals}
In Example III, we explore the performance of \pkg{R-INLA} using real data -- a publicly available dataset on mallard duck abundance in Switzerland in 2002.  By employing real data, we hoped to evaluate (1) the performance of \pkg{R-INLA} using data that were not predictable by design and (2) the practical outcome of not being able to specify site-survey covariates in \pkg{R-INLA}.  The dataset is available as a demonstration dataset in \pkg{unmarked}, so we compared the performance of \pkg{R-INLA} with \pkg{unmarked} in Example III.

\subsection[Analysis with unmarked]{Analysis with unmarked}
The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection.  Both sampling intensity and sampling date were site-survey covariates.

\begin{Code}
# get real mallard data from unmarked package
data(mallard)
# format unmarked data
mallard.y.unmk <- mallard.y
mallard.site.unmk <- mallard.site
mallard.obs.unmk <- mallard.obs
mallardUMF <- unmarkedFramePCount(mallard.y.unmk, siteCovs = mallard.site.unmk,
                                  obsCovs = mallard.obs.unmk)
# run unmarked model
mallard.out.unmk <- pcount(~ ivel+ date + I(date^2) ~ length + elev + forest,
                           mixture = "NB", mallardUMF, K=30)
summary(mallard.out.unmk)
\end{Code}

\subsection[Analysis with R-INLA]{Analysis with R-INLA}
The \pkg{R-INLA} model used a similar model structure.  The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead, which were attained using the rowMeans() function.

\begin{Code}
# format inla data
mallard.y.inla <- mallard.y
mallard.length.inla <- mallard.site[,2]
mallard.elev.inla <- mallard.site[,1]
mallard.forest.inla <- mallard.site[,3]
mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # average per site
mallard.ivel.inla[is.na(mallard.ivel.inla)] <- mean(mallard.ivel.inla, na.rm=T)
mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # average per site
mallard.date2.inla <- mallard.date.inla^2
# make inla.mdata object
counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, mallard.length.inla,
                                    mallard.elev.inla, mallard.forest.inla)
# run inla model
mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
                           mallard.date.inla + mallard.date2.inla,
         data = list(counts.and.count.covs = counts.and.count.covs,
                     mallard.ivel.inla = mallard.ivel.inla,
                     mallard.date.inla = mallard.date.inla,
                     mallard.date2.inla = mallard.date2.inla),
         family = "nmixnb",
         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
                              prec.intercept = 0.01),
         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
                                          theta2 = list(param = c(0, 0.01)),
                                          theta3 = list(param = c(0, 0.01)),
                                          theta4 = list(param = c(0, 0.01)),
                                          theta5 = list(prior = "flat",
                                                        param = numeric()))))
summary(mallard.out.inla)
\end{Code}

The 95\% credible intervals from the \pkg{R-INLA} model overlapped considerably with the 95\% confidence intervals from \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig. \ref{fig:fig3}). Also, the parameters with values significantly different from zero were the same regardless of technique (Fig. \ref{fig:fig3}).  Using both techniques, detection decreased linearly as the season progressed and abundance decreased as linear functions of increased forest cover and elevation gain.  Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \pkg{unmarked} and site-averaged detection covariates were used for \pkg{R-INLA}. Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would lead to similar results regardless of which covariate is used.  Conclusions may have been different with a different real dataset, where detection covariates had very strong effects.

%% figure 3
\begin{figure}
  \includegraphics[width=\linewidth]{fig3.jpeg}
  \caption{Estimates and 95\% confidence intervals from \pkg{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \pkg{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.  The \pkg{unmarked} model included site-survey covariates for survey effort and survey date, while the \pkg{R-INLA} model included site-averaged versions.  A value of zero is depicted by the vertical gray line.}
  \label{fig:fig3}
\end{figure}
%% end figure 3

%% discussion
\section[Discussion]{Discussion}
The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.

Results showed that \pkg{R-INLA} can be a complementary tool in the wildlife biologist's analytical tool kit.  Strengths of \pkg{R-INLA} include Bayesian inference, based on highly accurate approximations of posterior distributions, which are derived over 300 times faster than MCMC methods, where models are specified using a syntax that should be familiar to R users, and where data are formatted in a straightforward way with relatively few lines of code.  The straightforward model syntax and data format could help remove  barriers to adoption of N-mixture models for biologists who are not committed to learning the \proglang{BUGS} syntax.  The substantial decrease in computation time should allow a wider variety of model and variable selection techniques (e.g., cross validation and model averageing), ones that are not commonly used in an MCMC context due to practical issues related to computing time \citep{Kery_Schaub_2011}.

Limitations of \pkg{R-INLA} are mainly related to the more restricted set of N-mixture models that can be specified.  Of the approaches explored here, \proglang{BUGS}-based approaches allow users ultimate flexibility in specifying models. For example, with \proglang{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}.  In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-binomial and negative binomial-binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.  In cases where site-survey covariates are particularly strong and not controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig. \ref{fig:fig2}).

When compared to \pkg{unmarked}, the \pkg{R-INLA} approach is similar in regards to model syntax and data format.  The approaches are also similar in that both yield results much faster than MCMC.  The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, possibly due to the different approach used to compute model likelihoods (see Appendix).  They also differ in that \pkg{unmarked} can accomodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.

In conclusion, \proglang{JAGS} (and \proglang{WinBUGS}, \proglang{OpenBUGS}), \pkg{unmarked}, and \pkg{R-INLA} all allow users to employ N-mixture models for estimating wildlife abundance while accounting for imperfect detection.  Each method has its strengths and limitations.  \pkg{R-INLA} appears to be an attractive option when survey level covariates are not essential, familiar model syntax and data format are desired, fast computing time is needed, and Bayesian inference is preferred.

%% acknowledgements
\section*{Acknowledgments}
We thank the following individuals for commenting on previous drafts of this manuscript.

%% references
\bibliography{bibliog}

%% appendix
\appendix
\section{Recursive computations of the "nmix" likelihood}
The likelihood for the simplest case is

\begin{displaymath}
    \text{Probability}(y) = \sum_{n = y}^{\infty}
    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
\end{displaymath}

where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution
with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$. Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we
can easily evaluate it using a recursive algorithm that is both faster and more numerical stable. The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$. We leave it to the reader to derive these straight forward extentions.

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,

\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}

and

\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}

and then also for the Poisson-binomial product

\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}

If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as

\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}

The log-liklihood can then evaluated using the following simple \proglang{R}-code

\begin{verbatim}
fac = 1; ff = lambda * (1-p)
for (i in (n.max - y):1) fac = 1 + fac * ff / i
log.L = dpois(y, lambda, log=TRUE) +
        dbinom(y, y, p, log=TRUE) + log(fac)
\end{verbatim}

Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is
small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.

\end{document}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-04-20 meehan_et_al_inla_nmix_jss/meehan_et_al_inla_nmix_jss.R

@@ -1,0 +1,673 @@
+# code for Meehan, Michel, and Rue #############################################
+# demonstration of R-INLA for analyzing N-mixture models
+# 04/19/2017
+# ##############################################################################
+
+
+
+
+
+# directories, libraries, and helpers ##########################################
+setwd("~/GitHub/Quantitative_Metrics/INLA_NMix")
+library(runjags)
+library(INLA)
+library(unmarked)
+library(ggplot2)
+# multiplot function
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+  plots <- c(list(...), plotlist)
+  numPlots = length(plots)
+  if (is.null(layout)) {
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                     ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+  if (numPlots==1) {
+    print(plots[[1]])
+  } else {
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+    for (i in 1:numPlots) {
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+# plotting theme
+theme_acbs <- function (base_size = 11, base_family = "") {
+  theme_grey(base_size = base_size, base_family = base_family) %+replace%
+    theme(panel.background = element_rect(fill = "white", colour = NA),
+          panel.border = element_rect(fill = NA, colour = "grey20"),
+          panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          axis.text.x = element_text(size = rel(0.9), angle = 0),
+          axis.text.y = element_text(size = rel(0.9), angle = 0),
+          strip.background = element_rect(fill = "grey85", colour = "grey20"),
+          legend.key = element_rect(fill = "white", colour = NA),
+          complete = TRUE)
+}
+# data generating function
[…]

### 2017-04-20 meehan_et_al_inla_nmix_jss/article.tex
Small fix in my name
More fine tuning

@@
 %% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author{Timothy D. Meehan\\National Audubon Society \And
-        Nicole L. Michel\\National Audubon Society \And
+\author{Timothy D. Meehan\\National\\Audubon Society \And
+        Nicole L. Michel\\National\\Audubon Society \And
         H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
@@
 
-$$N_i \sim Poisson(\lambda) \qquad and \qquad  Y_{i,j} | N_i \sim binomial(N_i, p).$$
+$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{binomial}(N_i, p).$$
 
-$\lambda$ is commonly modeled as a log-linear function of site covariates, as $log(\lambda_i) = \beta_0 + \beta_1 * x_i$.
-Similarly, $p$ is commonly modeled as $logit(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.
+$\lambda$ is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1 * x_i$.
+Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.
 
@@
 In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
-A simple specification of a multiple-year model is $N_{i,k} \sim Poisson(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim binomial(N_{i,k}, p_{i,k})$.
+A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$.
 Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.
@@
 \begin{Code}
-data4sim <- function(n.sites = 72,   # number study sites
-                     n.surveys = 3,  # short term replicates
-                     n.years = 9,    # number years
-                     lam.b0 = 2.0,   # intercept for log lambda
-                     lam.b1 = 2.0,   # slope for log lambda, covariate 1
-                     lam.b2 = -3.0,  # slope for log lambda, covariate 2
-                     lam.b4 = 1.0,   # slope for log lambda, year
-                     p.b0 = 1.0,     # intercept for logit p
-                     p.b1 = -2.0,    # slope for logit p, covariate 1
-                     p.b3 = 1.0,     # slope for logit p covariate 3
-                     disp.size = 3.0 # size of the overdisperison
-                     ){
-  # setup
-  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
-  N.tr <- array(dim = c(n.sites, n.years))           # array for true abund
-  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex1 counts
-  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # array for ex2 counts
-  # abundance covariate values
-  x.lam.1 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
-                   dim=c(n.sites, n.years))
-  x.lam.2 <- array(as.numeric(scale(runif(n=n.sites, -0.5, 0.5), scale=F)),
-                   dim=c(n.sites, n.years))
-  yrs <- 1:n.years; yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
-  yr <- array(rep(yrs, each=n.sites), dim=c(n.sites, n.years))
-  # fill abundance array
-  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
-  for(i in 1:n.sites){
-    for(k in 1:n.years){
-    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
-  }}
-  # detection covariate values
-  x.p.1 <- array(x.lam.1[,1], dim=c(n.sites,n.surveys,n.years))
-  x.p.3 <- array(as.numeric(scale(runif(n=n.sites*n.surveys*n.years, -0.5, 0.5),
-                                  scale=F)), dim=c(n.sites,n.surveys,n.years))
-  # average x.p.3 per site-year
-  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
-  out1 <- c()
-  for(k in 1:n.years){
-   chunk1 <- x.p.3.mean[,k]
-   chunk2 <- rep(chunk1, n.surveys)
-   out1 <- c(out1, chunk2)
-  }
-  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
-  # fill count array with site-yr x.p.3
-  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
-  for (i in 1:n.sites){
-    for (k in 1:n.years){
-      for (j in 1:n.surveys){
-        y1[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr1[i,j,k])
-  }}}
-  # fill count array with site-surv-yr x.p.3
-  p.tr2 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3)
-  for (i in 1:n.sites){
-    for (k in 1:n.years){
-      for (j in 1:n.surveys){
-        y2[i,j,k] <- rbinom(1, size=N.tr[i,k], prob=p.tr2[i,j,k])
-  }}}
-  # return data
-  return(list(n.sites=n.sites, n.surveys=n.surveys, n.years=n.years,
-              x.p.1=x.p.1[,1,1], x.p.3=x.p.3, x.p.3.mean=x.p.3.mean,
-              x.p.3.arr=x.p.3.arr, x.lam.1=x.lam.1[,1],
-              x.lam.2=x.lam.2[,1], yr=yr[1,], y1=y1, y2=y2,
-              lam.tr=lam.tr, N.tr=N.tr))
-} #end function
-
-# create reproducible datasets for exercises 1 and 2
-set.seed(12345)
-sim.data <- data4sim()
+R> data4sim <- function(n.sites = 72,   # number study sites
++                       n.surveys = 3,  # short term replicates
++                       n.years = 9,    # number years
++                       lam.b0 = 2.0,   # intercept for log lambda
++                       lam.b1 = 2.0,   # slope for log lambda cov 1
++                       lam.b2 = -3.0,  # slope for log lambda cov 2
++                       lam.b4 = 1.0,   # slope for log lambda year
++                       p.b0 = 1.0,     # intercept for logit p
++                       p.b1 = -2.0,    # slope for logit p cov 1
++                       p.b3 = 1.0,     # slope for logit p cov 3
++                       disp.size = 3.0 # size of overdispersion
++                       ){
++  # setup
++  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
++  N.tr <- array(dim = c(n.sites, n.years))           # for abund
++  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
++  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
++  # abundance covariate values
++  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
++                   scale = F)), dim = c(n.sites, n.years))
++  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
++                   scale = F)), dim = c(n.sites, n.years))
++  yrs <- 1:n.years
++  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
++  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
++  # fill abundance array
++  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
++  for(i in 1:n.sites){
++    for(k in 1:n.years){
++    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
++  }}
++  # detection covariate values
++  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
++  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
++                                        -0.5, 0.5), scale=F)), 
++                 dim=c(n.sites, n.surveys, n.years))
++  # average x.p.3 per site-year
++  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
++  out1 <- c()
++  for(k in 1:n.years){
++   chunk1 <- x.p.3.mean[,k]
++   chunk2 <- rep(chunk1, n.surveys)
++   out1 <- c(out1, chunk2)
++  }
++  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
++  # fill count array with site-yr x.p.3
++  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
++  for (i in 1:n.sites){
++    for (k in 1:n.years){
++      for (j in 1:n.surveys){
++        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
++  }}}
++  # fill count array with site-surv-yr x.p.3
++  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
++  for (i in 1:n.sites){
++    for (k in 1:n.years){
++      for (j in 1:n.surveys){
++        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
++  }}}
++  # return data
++  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
++              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
++              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
++              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
++              lam.tr = lam.tr, N.tr = N.tr))
++  } #end function
+R> set.seed(12345) # make results reproducible
+R> sim.data <- data4sim() # create datasets
 \end{Code}
@@
 \begin{Code}
-jags.model.string <- "
-model {
-  # priors
-  intP ~ dnorm(0, 0.01)       # detection intercept
-  bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
-  bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
-  intLam ~ dnorm(0, 0.01)     # lambda intercept
-  bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
-  bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
-  bYr ~ dnorm(0, 0.01)        # year effect
-  overDisEst ~ dunif(0, 5)    # overdispersion size
-  # abundance component
-  for (k in 1:nYears){
-    for (i in 1:nSites){
-      N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
-      prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
-      log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
-                          (bYr * yr[k])
-  # detection component
-      for (j in 1:nSurveys){
-        y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
-        p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
-        lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
-      } # close j loop
-    } # close i loop
-  } # close k loop
-} # close model loop
-"
-}
+R> jags.model.string <- "
++    model {
++    # priors
++    intP ~ dnorm(0, 0.01)       # detection intercept
++    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
++    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
++    intLam ~ dnorm(0, 0.01)     # lambda intercept
++    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
++    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
++    bYr ~ dnorm(0, 0.01)        # year effect
++    overDisEst ~ dunif(0, 5)    # overdispersion size
++    # abundance component
++    for (k in 1:nYears){
++      for (i in 1:nSites){
++        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
++        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
++        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
++                          (bYr * yr[k])
++    # detection component
++        for (j in 1:nSurveys){
++          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
++          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
++          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
++        } # close j loop
++      } # close i loop
++    } # close k loop
++  } # close model loop
++  "
 \end{Code}
@@
 \begin{Code}
-# parameters to monitor
-params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
-            "bYr", "overDisEst")
-# jags data
-jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
-             x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
-             x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
-             nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
-             nYears = sim.data$n.years)
-# initial values
-N.init <- sim.data$y1 # initial count values
-N.init[is.na(N.init)] <- 1 # clean up NA's
-N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
-inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
-                     intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
-                     bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
-                     bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
-                     overDisEst = runif(1, 0.5, 2.5))
+
+R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
++              "bYr", "overDisEst") # parameters to monitor
+R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
++               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
++               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
++               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
++               nYears = sim.data$n.years)
+R> N.init <- sim.data$y1 # initial count values
+R> N.init[is.na(N.init)] <- 1 # clean up NA's
+R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
+R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
++                      intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
++                      bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
++                      bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
++                      overDisEst = runif(1, 0.5, 2.5))
 \end{Code}
@@
 \begin{Code}
-# set run parameters
-nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10
-# run jags
-out.jags <- run.jags(model = jags.model.string, data = jags.data,
-                     monitor = params, n.chains = nc, inits = inits,
-                     burnin = nb, adapt = na, sample = ni, thin = nt,
-                     modules = "glm on", method = "parallel")
-summary(out.jags)
+R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # set run parameters
+R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
++                       monitor = params, n.chains = nc, inits = inits,
++                       burnin = nb, adapt = na, sample = ni, thin = nt,
++                       modules = "glm on", method = "parallel")
 \end{Code}
@@
 \begin{Code}
-# format count data
-y.unmk <- sim.data$y1[ , , 1]
-for(i in 2:sim.data$n.years){
-  y.chunk <- sim.data$y1[ , , i]
-  y.unmk <- rbind(y.unmk, y.chunk)
-}
-# format covariates
-x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
-x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
-yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
-x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
-x.p.3.unmk <- c(sim.data$x.p.3.mean)
-site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
-                             x.p.1.unmk, x.p.3.unmk)
-# make unmarked pcount frame
-unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
+R> y.unmk <- sim.data$y1[ , , 1]
+R> for(i in 2:sim.data$n.years){
++    y.chunk <- sim.data$y1[ , , i]
++    y.unmk <- rbind(y.unmk, y.chunk)
++  } 
+R> x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
+R> x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
+R> yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
+R> x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
+R> x.p.3.unmk <- c(sim.data$x.p.3.mean)
+R> site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
++                               x.p.1.unmk, x.p.3.unmk)
+R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
 \end{Code}
@@
 \begin{Code}
-# run unmk model
-out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
-                   ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
-                    data = unmk.data, mixture = "NB")
-summary(out.unmk)
+R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
++                     ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
++                     data = unmk.data, mixture = "NB")
 \end{Code}
@@
 \begin{Code}
-# format count data
-y.inla <- sim.data$y1[ , , 1]
-for(i in 2:sim.data$n.years){
-  y.chunk <- sim.data$y1[ , , i]
-  y.inla <- rbind(y.inla, y.chunk)
-}
-# format covariates
-x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
-x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
-yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
-x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
-x.p.3.inla <- c(sim.data$x.p.3.mean)
-# make inla.mdata object
-counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, x.lam.2.inla, yr.inla)
+
+R> y.inla <- sim.data$y1[ , , 1]
+R> for(i in 2:sim.data$n.years){
++    y.chunk <- sim.data$y1[ , , i]
++    y.inla <- rbind(y.inla, y.chunk)
++  }
+R> x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
+R> x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
+R> yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
+R> x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
+R> x.p.3.inla <- c(sim.data$x.p.3.mean)
+R> counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, 
++                                      x.lam.2.inla, yr.inla)
 \end{Code}
@@
 \begin{Code}
-# run inla model
-out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
-         data = list(counts.and.count.covs = counts.and.count.covs,
-                     x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
-         family = "nmixnb",
-         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
-                              prec.intercept = 0.01),
-         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
-                                          theta2 = list(param = c(0, 0.01)),
-                                          theta3 = list(param = c(0, 0.01)),
-                                          theta4 = list(param = c(0, 0.01)),
-                                          theta5 = list(prior = "flat",
-                                                        param = numeric()))))
-summary(out.inla)
+R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
++           data = list(counts.and.count.covs = counts.and.count.covs,
++                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
++           family = "nmixnb",
++           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
++                                prec.intercept = 0.01),
++           control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
++                                              theta2 = list(param = c(0, 0.01)),
++                                              theta3 = list(param = c(0, 0.01)),
++                                              theta4 = list(param = c(0, 0.01)),
++                                              theta5 = list(prior = "flat",
++                                                            param = numeric()))))
 \end{Code}
@@
 \begin{Code}
-# get real mallard data from unmarked package
-data(mallard)
-# format unmarked data
-mallard.y.unmk <- mallard.y
-mallard.site.unmk <- mallard.site
-mallard.obs.unmk <- mallard.obs
-mallardUMF <- unmarkedFramePCount(mallard.y.unmk, siteCovs = mallard.site.unmk,
-                                  obsCovs = mallard.obs.unmk)
-# run unmarked model
-mallard.out.unmk <- pcount(~ ivel+ date + I(date^2) ~ length + elev + forest,
-                           mixture = "NB", mallardUMF, K=30)
-summary(mallard.out.unmk)
+R> data("mallard") # get mallard data from unmarked package
+R> mallard.y.unmk <- mallard.y # format unmarked data
+R> mallard.site.unmk <- mallard.site
+R> mallard.obs.unmk <- mallard.obs
+R> mallardUMF <- unmarkedFramePCount(mallard.y.unmk, 
++                                    siteCovs = mallard.site.unmk,
++                                    obsCovs = mallard.obs.unmk)
+R> mallard.out.unmk <- pcount(~ ivel+ date + I(date^2)
++                             ~ length + elev + forest,
++                             mixture = "NB", mallardUMF, K=30) 
 \end{Code}
@@
 \begin{Code}
-# format inla data
-mallard.y.inla <- mallard.y
-mallard.length.inla <- mallard.site[,2]
-mallard.elev.inla <- mallard.site[,1]
-mallard.forest.inla <- mallard.site[,3]
-mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # average per site
-mallard.ivel.inla[is.na(mallard.ivel.inla)] <- mean(mallard.ivel.inla, na.rm=T)
-mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # average per site
-mallard.date2.inla <- mallard.date.inla^2
-# make inla.mdata object
-counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, mallard.length.inla,
-                                    mallard.elev.inla, mallard.forest.inla)
-# run inla model
-mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
-                           mallard.date.inla + mallard.date2.inla,
-         data = list(counts.and.count.covs = counts.and.count.covs,
-                     mallard.ivel.inla = mallard.ivel.inla,
-                     mallard.date.inla = mallard.date.inla,
-                     mallard.date2.inla = mallard.date2.inla),
-         family = "nmixnb",
-         control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
-                              prec.intercept = 0.01),
-         control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
-                                          theta2 = list(param = c(0, 0.01)),
-                                          theta3 = list(param = c(0, 0.01)),
-                                          theta4 = list(param = c(0, 0.01)),
-                                          theta5 = list(prior = "flat",
-                                                        param = numeric()))))
-summary(mallard.out.inla)
+R> mallard.y.inla <- mallard.y # format inla data
+R> mallard.length.inla <- mallard.site[,2]
+R> mallard.elev.inla <- mallard.site[,1]
+R> mallard.forest.inla <- mallard.site[,3]
+R> mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # site average
+R> mallard.ivel.inla[is.na(mallard.ivel.inla)] <- 
++                                       mean(mallard.ivel.inla, na.rm = T)
+R> mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # site average
+R> mallard.date2.inla <- mallard.date.inla^2
+R> counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, 
++                                      mallard.length.inla,
++                                      mallard.elev.inla, 
++                                      mallard.forest.inla)
+R> mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
++                           mallard.date.inla + mallard.date2.inla,
++      data = list(counts.and.count.covs = counts.and.count.covs,
++                           mallard.ivel.inla = mallard.ivel.inla,
++                           mallard.date.inla = mallard.date.inla,
++                           mallard.date2.inla = mallard.date2.inla),
++      family = "nmixnb",
++      control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
++                               prec.intercept = 0.01),
++      control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
++                                    theta2 = list(param = c(0, 0.01)),
++                                    theta3 = list(param = c(0, 0.01)),
++                                    theta4 = list(param = c(0, 0.01)),
++                                    theta5 = list(prior = "flat",
++                                                  param = numeric()))))
 \end{Code}
@@
 The likelihood for the simplest case is
-
 \begin{displaymath}
@@
 \end{displaymath}
-
 where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution
@@
 The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$.
-We leave it to the reader to derive these straight forward extentions.
+We leave it to the reader to derive these straight forward extensions.
 
 The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
-
 \begin{displaymath}
@@
 \end{displaymath}
-
 and
-
 \begin{displaymath}
@@
 \end{displaymath}
-
 and then also for the Poisson-binomial product
-
 \begin{displaymath}
@@
 \end{displaymath}
-
 If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
-
 \begin{eqnarray}
@@
 \end{eqnarray}
-
 The log-liklihood can then evaluated using the following simple \proglang{R}-code
-
-\begin{verbatim}
-fac = 1; ff = lambda * (1-p)
-for (i in (n.max - y):1) fac = 1 + fac * ff / i
-log.L = dpois(y, lambda, log=TRUE) +
-        dbinom(y, y, p, log=TRUE) + log(fac)
-\end{verbatim}
-
-Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is
-small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
+\begin{Code}
+R> fac <- 1; ff <- lambda * (1-p)
+R> for(i in (n.max - y):1){
++    fac <- 1 + fac * ff / i
++  } 
+R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
++           log(fac)
+\end{Code}
+Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.
+Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
 

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
More fine tuning
Fixed figures

@@
 
-%% for pretty printing
 %% comma-separated
@@
 }
-\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \proglang{R}, \proglang{BUGS}}
+\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \pkg{unmarked}, \proglang{R}, \proglang{JAGS}}
 %% without formatting
@@
 R> data4sim <- function(n.sites = 72,   # number study sites
-+                       n.surveys = 3,  # short term replicates
++                       n.surveys = 3,  # number replicate surveys
 +                       n.years = 9,    # number years
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{fig1.jpeg}
+  \includegraphics[width=\linewidth]{fig1.png}
   \caption{Marginal posteriors from \proglang{JAGS} (solid gray lines) and \pkg{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \pkg{unmarked}.
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{fig2.jpeg}
+  \includegraphics[width=\linewidth]{fig2.png}
   \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{fig3.jpeg}
+  \includegraphics[width=\linewidth]{fig3.png}
   \caption{Estimates and 95\% confidence intervals from \pkg{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \pkg{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.
@@
 \end{eqnarray}
-The log-liklihood can then evaluated using the following simple \proglang{R}-code
+The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
 \begin{Code}
@@
 \end{Code}
-Since this evaluation is recursive in decreasing $n$, we have to chose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.
+Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.
 Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.

### 2017-04-21 meehan_et_al_inla_nmix_jss/meehan_et_al_inla_nmix_jss.R

@@ -306,11 +306,12 @@ d5 <- all.out.tab[col1, 6]
 d6 <- (c(d3,d4,d5))
 d7 <- data.frame(y=c(0,0,0), x=d6)
 p1 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) +
-  geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) + xlab("log(Lambda)") +
+  geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) +
+  xlab("log(lambda) intercept") +
   ylab(" ") + scale_y_continuous(breaks=c(3,6,9)) + theme_acbs() +
   geom_vline(xintercept=(2), col="gray20", lty=1) +
-geom_line(data=d7, aes(y=y, x=x)) +
-geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
+  geom_line(data=d7, aes(y=y, x=x)) + theme(axis.title.x = element_text(size=9)) +
+  geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.1.lam
 d1 <- as.data.frame(out.inla$marginals.hyperpar$`beta[2] for NMix observations`)
 d2 <- as.data.frame(density(jags.df$bCov1Lam)[c(1,2)])
@@ -322,10 +323,11 @@ d6 <- c(d3,d4,d5)
 d7 <- data.frame(y=c(0,0,0), x=d6)
 p2 <- ggplot(data=d2, aes(x=x, y=y)) + geom_path(col="gray60", lty=1) +
   geom_path(data=d1, aes(x=x, y=y), col="black", lty=2 ) +
-  xlab("Abundance covariate 1")  + ylab("Posterior density") + theme_acbs() +
+  xlab("log(lambda) covariate 1")  +
+  ylab("Posterior density") + theme_acbs() +
   geom_vline(xintercept=2, col="gray20", lty=1) +
-geom_line(data=d7, aes(y=y, x=x)) +
-geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
+  geom_line(data=d7, aes(y=y, x=x)) + theme(axis.title.x = element_text(size=9)) +
+  geom_point(data=d7, aes(y=y, x=x), col="black", shape=c(32,19,32))
 # make density for x.2.lam
[…]

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
Ready to send off after you look it over
@hrue , @nicolelmichel
Hi Haavard, Nicole.  I have fine tuned the R-INLA nMix manuscript.  In
tuning, I hope I haven't messed anything up.  If you have time, please
look it over one more time.  After I get your approval, I will send it
off for review.  The manuscript is now in LaTeX format.  The latest
version is also compiled as a PDF for convenience. Feel free to work
with either one. The LaTeX format follow the JSS style.  Seems like this
is an appropriate outlet for this manuscript.  Let me know if you have
other preferences. Also, let me know if yo

@@
 Abundance estimates can be confounded by imperfect detection during wildlife surveys.
-Development of N-mixture models has facilitated quantification of detectability and has improved our ability to produce accurate abundance estimates.
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo methods and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference.
-Our results suggest that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) survey level covariates are not essential, (2) familiar model syntax and data format are desired (relative to other R packages), (3) fast computing time is necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
-
+N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased.
+The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference.
+We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
 }
-\Keywords{wildlife, abundance, detection, N-mixture models, \pkg{R-INLA}, \pkg{unmarked}, \proglang{R}, \proglang{JAGS}}
+\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
 %% without formatting
@@
 One common method for estimating animal abundance is direct counts \citep{Pollock_Nichols_Simons_Farnsworth_Bailey_Sauer_2002}.
-Efforts to obtain accurate abundance estimates via direct counts can be hindered by both the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure.
+Efforts to obtain accurate abundance estimates via direct counts can be hindered by the cryptic nature of many wildlife species and by other factors such as observer expertise, weather, and habitat structure \citep{Denes_Silveira_Beissinger_2015}.
 The lack of perfect detection in wildlife surveys is common, and can cause abundance to be greatly underestimated \citep{Wenger_Freeman_2008, Joseph_Elkin_Martin_Possingham_2009}.
@@
 In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}.
-One such sampling scheme, termed a "metapopulation design" \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.
-If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
-This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Denes_Silveira_Beissinger_2015}.
+One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.
+If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
+This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.
 
@@
 
-This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for annually-breeding wildlife species \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
+This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
 In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
@@
 Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.
-There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.
+There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, a zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.
 Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
-We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for a nice overview.
+We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.
 
@@
 However, there are practical issues that sometimes act as barriers to adoption.
-Much of the development of N-mixture models, and many of the examples in the wildlife literature, implement models using Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
-These are extremely powerful and flexible platforms for analyzing Bayesian models, but they come with a few important challenges.
+Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
+These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.
 First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax.
@@
 The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.
-However, it comes at a cost -- the loss of the intuitive inferential framework associated with Bayesian analysis.
+However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.
 
@@
 INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC.
-Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in R.
+Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}.
 Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low.
@@
 Rather, survey level covariates of detection must be averaged to the site or site-year level.
-Using an averaged detection covariate does allow accounting for site-level differences in survey conditions should they occur.
-However, in the process of averaging, important information related to detection is discarded, which could lead to biased detection and abundance estimates under certain conditions.
+Using an averaged detection covariate does allow accounting for site-level differences in survey conditions, should they occur.
+However, in the process of averaging, information related to detection within a site or site-year combination is discarded, which could lead to biased detection and abundance estimates under certain conditions.
 
@@
 \subsection[Simulated data]{Simulated data}
-The data simulated for this analysis were intended to represent a typical wildlife study.
+The data simulated for this analysis were intended to represent a typical wildlife population study.
 To put the simulation in context, consider an effort to estimate the abundance of a bird species in a national park, within which are located 72 study sites.
-At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season when birds are likely to be singing.
-In order to estimate a trend in abundance over time, these repeated surveys are conducted over a 9-year period.
+At each site, 3 replicate surveys are conducted within 6 weeks, during the peak of the breeding season, when birds are likely to be singing.
+In order to estimate a trend in abundance over time, clusters of repeated surveys are conducted each breeding season over a 9-year period.
 
@@
 Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model.
-The specific model parameters and their simulated values were: p.b0 = 1.0 ($p$ intercept), p.b1 = -2.0 (effect of $p$ covariate 1), p.b3 = 1.0 (effect of $p$ covariate 3), lam.b0 = 2.0 ($\lambda$ intercept), lam.b1 = 2.0 (effect of $\lambda$ covariate 1), lam.b2 = -3.0 (effect of $\lambda$ covariate 2), lam.b4 = 1.0 (effect of year on $\lambda$), disp.size = 3.0 (size of the overdispersion parameter).
+The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter).
 All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.
@@
 Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate.
-This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three packages when the specified model was reasonably close to the data generating process.
+This exercise was intended to demonstrate the differences and similarities in use, computation time, and estimation results across the three methods when the specified model was reasonably close to the data generating process.
 
@@
 +      for (i in 1:nSites){
-+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial specification
-+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k]) # overdispersion effect
-+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + (bCov2Lam * x.lam.2[i]) +
-+                          (bYr * yr[k])
++        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
++        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
++        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
++                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
 +    # detection component
@@
 \begin{Code}
-
 R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
@@
 +               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
-+                      bCov2Lam = rnorm(1,0,0.01), bCov1Lam = rnorm(1, 0, 0.01),
++               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
 +               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
@@
 Finally, we set the run parameters and started the MCMC process.
-Run parameters were chosen such that MCMC diagnostics indicated converged chains (Gelman-Rubin statistics < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).
+Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).
 Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}.
@@
 \begin{Code}
-R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # set run parameters
+R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
 R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
@@
 \ref{fig:fig1}).
-The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2410.
+The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2409.
 The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.
@@
 
-Maximum likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig.
+Maximum Likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig.
 \ref{fig:fig1})  The \pkg{unmarked} estimates were produced in approximately 77 seconds.
@@
 On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.
-On the right side of the formula is a 1 to specify a global intercept for $p$ and the two fixed covariates for $p$.
-Note that a wide range of random effects for $p$ could be added to the right side of the formula \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}.
+On the right side of the formula is a 1 to specify a global intercept for $p$ and the two covariates for $p$.
+Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}.
 The second argument describes the data, provided here as a list that corresponds with the model formula.
 Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture.
-The fourth and fifth arguments specify the priors for the two model components.
+The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components.
 In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available.
-Several other characteristics of the analysis can be modified using the call to inla().
+Several other characteristics of the analysis can be modified in the call to inla().
 See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.
@@
 +                                prec.intercept = 0.01),
-+           control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
++           control.family = list(hyper = 
++                                  list(theta1 = list(param = c(0, 0.01)),
 +                                       theta2 = list(param = c(0, 0.01)),
@@
 Mean parameter estimates from the \pkg{R-INLA} model are also close to, and not significantly different from, input values (Fig.
-\ref{fig:fig1})  Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig.
+\ref{fig:fig1}).
+Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig.
 \ref{fig:fig1}.
@@
 In demonstrating the use of \pkg{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.
-Similarly, model specification uses a straightforward extension of the standard syntax in R, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula.
-Data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.
+Similarly, model specification uses a straightforward extension of the standard syntax in \proglang{R}, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula.
+The data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.
 
@@
 Fig.
-\ref{fig:fig1} depicts marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and 95\% confidence intervals from \pkg{unmarked}.
-These results derive from data resulting from one random manifestation of the input values.
+\ref{fig:fig1} shows marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and estimates and 95\% confidence intervals from \pkg{unmarked}.
+These results derive from data from one random manifestation of the input values.
 Thus, we do not expect the posterior distributions for the estimates to be centered at the input values, which would be expected if the simulation was repeated many times.
-However, we do expect the input values to fall somewhere within the posterior distributions and 95\% confidence intervals, which is what occurs here.
+However, we do expect the input values to fall somewhere within the posterior distributions and 95\% confidence limits, which is what occurs here.
 Fig.
@@
 Thus, \pkg{R-INLA} was approximately 10 times faster than \pkg{unmarked} and 300 times faster than \proglang{JAGS}.
-This was the case despite the fact that \pkg{unmarked} uses ML algorithms and the \proglang{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core.
+This was the case despite the fact that \pkg{unmarked} produces ML estimates and the \proglang{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core.
 If parallel computing had not been used with \proglang{JAGS}, processing the \proglang{JAGS} model would have taken approximately twice as long.
@@
 Its utility depends on the degree to which the data generating process can be captured accurately in model specification.
-However, as mentioned above, certain models can not be specified using \pkg{R-INLA}.
+However, as mentioned above, certain N-mixture models can not be specified using \pkg{R-INLA}.
 For the data in Example I, the count matrix was produced using a detection covariate that was averaged to the site-year level.
-This averaged covariate was subsequently used in the analysis.
-But what happens when the site-survey-year covariate is an important component of the data generating process, and it can't be specified in the model in this form?
+This averaged covariate was subsequently specified in the model.
+But what happens when the site-survey-year covariate is an important component of the data generating process, and it can't be entered into the model in this form?
 This is the question explored in Example II.
@@
 One model incorporated an averaged site-survey x.p.3.mean as a covariate, such as in 2.1.
-The other model incorporated the site-survey-year covariate instead.
+The other model incorporated the full site-survey-year covariate instead.
 In each iteration we randomly varied the size of the effect for x.p.3.
-We suspected that the simpler model, with x.p.3.mean, would yield biased estimates when the effect of x.p.3 was relatively large, and unbiased estimates when the effect was relatively small.
+We expected that the simpler model, with x.p.3.mean, would yield biased estimates when the effect of x.p.3 was relatively large, and unbiased estimates when the effect was relatively small.
 
 \subsection[Analysis with JAGS]{Analysis with JAGS}
-Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (after 1000 adaptive iterations and 1000 burn-in iterations) during each simulation.
+Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation.
 This number is not sufficient for drawing inference from marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means.
-In each of the 50 iterations the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3.
+In each of the 50 iterations, the size of the effect for x.p.3 was drawn from a uniform distribution that ranged from -3 to 3.
 The results of the simulations are depicted in Fig.
@@
 \subsection[Example II summary]{Example II summary}
-As expected, biases in estimates of abundance (abundance intercept) and the effects of abundance covariates (specifically, abundance covariate 1) increased with the effect strength of the site-survey-year covariate for $p$.
-When the effect size was small, with an absolute value less than 1, the bias was minimal.
-When the effect size was large, with an absolute value greater than 2, the bias was large.
+As expected, biases in estimates of average abundance (log(lambda) intercept) and the effects of abundance covariates (log(lambda) covariate 1) increased with the effect strength of the site-survey-year covariate for $p$ (x.p.3;  Fig.
+\ref{fig:fig2}).
+When the effect size was small, with an absolute value less than 1, the bias was negligible.
+When the effect size was large, with an absolute value greater than 2, the bias was considerable (Fig.
+\ref{fig:fig2}).
 When interpreting the effect size, bear in mind that x.p.3 ranged from -0.5 to 0.5.
@@
 By employing real data, we hoped to evaluate (1) the performance of \pkg{R-INLA} using data that were not predictable by design and (2) the practical outcome of not being able to specify site-survey covariates in \pkg{R-INLA}.
-The dataset is available as a demonstration dataset in \pkg{unmarked}, so we compared the performance of \pkg{R-INLA} with \pkg{unmarked} in Example III.
+The dataset is available as a demonstration dataset in \pkg{unmarked}, so we compared the performance of \pkg{R-INLA} with \pkg{unmarked}, using the demonstration model and analysis settings, in Example III.
 
@@
 The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection.
-Both sampling intensity and sampling date were site-survey covariates.
+Both sampling intensity and sampling date were site-survey, or survey level covariates.
 
@@
 The \pkg{R-INLA} model used a similar model structure.
-The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead, which were attained using the rowMeans() function.
+The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead.
+These averages were attained using the rowMeans() function.
 
@@
 
-The 95\% credible intervals from the \pkg{R-INLA} model overlapped considerably with the 95\% confidence intervals from \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
+The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
 \ref{fig:fig3}).
-Also, the parameters with values significantly different from zero were the same regardless of technique (Fig.
+Also, the parameters with estimates significantly different from zero were the same regardless of technique (Fig.
 \ref{fig:fig3}).
-Using both techniques, detection decreased linearly as the season progressed and abundance decreased as linear functions of increased forest cover and elevation gain.
+Using both techniques, detection decreased linearly as the season progressed, and abundance decreased as linear functions of increasing forest cover and elevation gain.
 Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \pkg{unmarked} and site-averaged detection covariates were used for \pkg{R-INLA}.
-Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would lead to similar results regardless of which covariate is used.
-Conclusions may have been different with a different real dataset, where detection covariates had very strong effects.
+Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would indicate that other parameter estimates were not substantially biased.
+These results may have been different with a different real dataset, where detection covariates had very strong effects.
 
@@
 The straightforward model syntax and data format could help remove  barriers to adoption of N-mixture models for biologists who are not committed to learning the \proglang{BUGS} syntax.
-The substantial decrease in computation time should allow a wider variety of model and variable selection techniques (e.g., cross validation and model averageing), ones that are not commonly used in an MCMC context due to practical issues related to computing time \citep{Kery_Schaub_2011}.
+The substantial decrease in computation time should facilitate use of a wider variety of model and variable selection techniques (e.g., cross validation and model averaging), ones that are not commonly used in an MCMC context due to practical issues related to computing time \citep{Kery_Schaub_2011}.
 
@@
 In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-binomial and negative binomial-binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.
-In cases where site-survey covariates are particularly strong and not controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig.
+In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig.
 \ref{fig:fig2}).
@@
 The approaches are also similar in that both yield results much faster than MCMC.
-The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, possibly due to the different approach used to compute model likelihoods (see Appendix).
-They also differ in that \pkg{unmarked} can accomodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
+The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, possibly due to the different approach used to compute model likelihoods (see Appendix A).
+They also differ in that \pkg{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
 
-In conclusion, \proglang{JAGS} (and \proglang{WinBUGS}, \proglang{OpenBUGS}), \pkg{unmarked}, and \pkg{R-INLA} all allow users to employ N-mixture models for estimating wildlife abundance while accounting for imperfect detection.
+In conclusion, \proglang{JAGS} (and \proglang{WinBUGS}, \proglang{OpenBUGS}), \pkg{unmarked}, and \pkg{R-INLA} all allow users to analyze N-mixture models for estimating wildlife abundance while accounting for imperfect detection.
 Each method has its strengths and limitations.
@@
 \end{displaymath}
-where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution
-with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
-Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we
-can easily evaluate it using a recursive algorithm that is both faster and more numerical stable.
+where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
+Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we can easily evaluate it using a recursive algorithm that is both faster and more numerical stable.
 The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$.

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
fixed code format to italic

@@
 
-\begin{Code}
+\begin{CodeInput}
 R> data4sim <- function(n.sites = 72,   # number study sites
@@
 R> sim.data <- data4sim() # create datasets
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> jags.model.string <- "
@@
 +  "
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
@@
 +               overDisEst = runif(1, 0.5, 2.5))
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
@@
 +                       modules = "glm on", method = "parallel")
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> y.unmk <- sim.data$y1[ , , 1]
@@
 R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
@@
 +                       data = unmk.data, mixture = "NB")
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 
@@
 +                                      x.lam.2.inla, yr.inla)
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
@@
 +                                                     param = numeric()))))
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> data("mallard") # get mallard data from unmarked package
@@
 +                               mixture = "NB", mallardUMF, K=30) 
-\end{Code}
+\end{CodeInput}
 
@@
 
-\begin{Code}
+\begin{CodeInput}
 R> mallard.y.inla <- mallard.y # format inla data
@@
 +                                                  param = numeric()))))
-\end{Code}
+\end{CodeInput}
 
@@
 The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
-\begin{Code}
+\begin{CodeInput}
 R> fac <- 1; ff <- lambda * (1-p)
@@
 +           log(fac)
-\end{Code}
+\end{CodeInput}
 Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
tweaks
incorporated haavard's comments

@@
         Nicole L. Michel\\National\\Audubon Society \And
-        H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
+        Haavard Rue\\King Abdulla University\\of Science and Technology}
 \title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}
@@
 %% comma-separated
-\Plainauthor{Timothy D. Meehan, Nicole L. Michel, H\"{a}vard Rue}
+\Plainauthor{Timothy D. Meehan, Nicole L. Michel, Haavard Rue}
 %% without formatting
@@
   \\
-  H\"{a}vard Rue\\
+  Haavard Rue\\
   King Abdulla University
   of Science and Technology\\
-  Thuwal, SAU\\
+  Thuwal, Saudi Arabia\\
   E-mail: \email{haavard.rue@kaust.edu.sa }
@@
 One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.
-If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
+If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
 This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.
@@
 Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture model \citep{Royle_Nichols_2003, Dodd_Dorazio_2004, Royle_2004, Kery_Royle_Schmid_2005}.
-A simple form of an N-mixture model, a binomial mixture model, describes observed counts $Y$ at site $i$ during survey $j$ as coming from a binomial distribution with parameters for abundance $N$ and detection probability $p$, where $N$ per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$.
+A simple form of an N-mixture model, a Binomial mixture model, describes observed counts $Y$ at site $i$ during survey $j$ as coming from a Binomial distribution with parameters for abundance $N$ and detection probability $p$, where $N$ per site is drawn from a Poisson distribution with an expected value $\lambda$.
 Specifically,
 
-$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{binomial}(N_i, p).$$
+$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{Binomial}(N_i, p).$$
 
-$\lambda$ is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1 * x_i$.
-Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.
+The mean is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1x_i$.
+Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1x_{i,j}$, a logit-linear function of site-survey covariates.
 
@@
 In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
-A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$.
+A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$.
 Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.
-There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, a zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.
-Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
+There are other variations of N-mixture models that accommodate overdispersed counts through use of a Negative Binomial distribution \citep{Kery_Royle_2010}, a Zero-Inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}.
+Yet other variations account for non-independent detection probabilities through use of a Beta-Binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
 We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.
@@
 However, there are practical issues that sometimes act as barriers to adoption.
-Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
+Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
 These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.
@@
 While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment.
-Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.
+Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, or may not occur at all.
 
@@
 
-Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} for \proglang{R}.
-The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009}.
+Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \proglang{R}.
+The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}.
 INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC.
@@
 All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}.
+See \url{https://r-inla.org} to connect with the community around the development of \pkg{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.
 
@@
 As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables.
-Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model.
+Overdispersion was generated and modeled using a Negative Binomial distribution for the count component of the model.
 The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter).
@@
 We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package.
-In defining the model, we specified a negative binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.
+In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.
 
@@
 The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance.
-This was followed by an argument identifying the appropriate unmarked frame, and the form of the mixture model, negative binomial-binomial.
+This was followed by an argument identifying the appropriate unmarked frame, and the form of the mixture model, Negative Binomial-Binomial.
 
@@
 The second argument describes the data, provided here as a list that corresponds with the model formula.
-Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture.
+Third is the likelihood family, which can take values of "nmix" for a Poisson-Binomial mixture and "nmixnb" for a Negative Binomial-Binomial mixture.
 The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components.
@@
 \section[Discussion]{Discussion}
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.
+The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.
 
@@
 For example, with \proglang{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}.
-In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-binomial and negative binomial-binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.
+In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-Binomial and Negative Binomial-Binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.
 In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig.
@@
 The approaches are also similar in that both yield results much faster than MCMC.
-The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, possibly due to the different approach used to compute model likelihoods (see Appendix A).
+The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, likely due to the different approach used to compute model likelihoods (see Appendix A).
 They also differ in that \pkg{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
@@
     \text{Probability}(y) = \sum_{n = y}^{\infty}
-    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
+    \text{Poisson}(n ; \lambda) \;\times\; \text{Binomial}(y;  n, p)
 \end{displaymath}
-where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
+where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{Binomial}(y; n, p)$ is the density for the Binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
 Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we can easily evaluate it using a recursive algorithm that is both faster and more numerical stable.
-The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$.
+The same idea is also applicable to the Negative Binomial case, and the case where we have replicated observations of the same $n$.
 We leave it to the reader to derive these straight forward extensions.
 
-The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
+The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
 \begin{displaymath}
@@
 \begin{displaymath}
-    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
+    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
 \end{displaymath}
-and then also for the Poisson-binomial product
+and then also for the Poisson-Binomial product
 \begin{displaymath}
-    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
+    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
     =
-    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
+    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
     \frac{\lambda}{n-y}(1-p).
@@
                        \text{Poisson}(n ; \lambda)\;
-                      \text{binomial}(y;  n, p) \nonumber\\
-                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
+                      \text{Binomial}(y;  n, p) \nonumber\\
+                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                        \Big\{ 1 + f_1 + f_1f_2 +
@@
                        \Big\} \nonumber\\
-                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
+                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                        \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
@@
 Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.
-Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
+Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
 

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
more tweaks

@@
         Nicole L. Michel\\National\\Audubon Society \And
-        Haavard Rue\\King Abdulla University\\of Science and Technology}
+        H{\aa}vard Rue\\King Abdulla University\\of Science and Technology}
 \title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}
@@
   \\
-  Haavard Rue\\
+  H{\aa}vard Rue\\
   King Abdulla University
@@
 One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.
-If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
+If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
 This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.
@@
 
-\subsection[Analysis with JAGS]{Analysis with JAGS}
+\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
 We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package.
@@
 
-\subsection[Analysis with unmarked]{Analysis with unmarked}
+\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
 Next, we prepared the simulated data for the \pkg{unmarked} analysis, which involved slight modification of the data in the sim.data object.
@@
 
-\subsection[Analysis with R-INLA]{Analysis with R-INLA}
+\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
 Next, we prepared data for \pkg{R-INLA}, which uses data formats similar to those used by \pkg{unmarked}.
@@
 
-\subsection[Analysis with JAGS]{Analysis with JAGS}
+\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
 Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation.
@@
 
-\subsection[Analysis with unmarked]{Analysis with unmarked}
+\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
 The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection.
@@
 
-\subsection[Analysis with R-INLA]{Analysis with R-INLA}
+\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
 The \pkg{R-INLA} model used a similar model structure.
@@
 \section*{Acknowledgments}
-We thank the following individuals for commenting on previous drafts of this manuscript.
+We thank Jerry Cole and Curtis Burkhalter for commenting on previous drafts of this manuscript.
 
@@
 \begin{eqnarray}
-    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
+    \text{Probability}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                        \text{Poisson}(n ; \lambda)\;

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
got Nicole's comments integrated

@@
 N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased.
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows Frequentist inference.
+The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows frequentist inference.
 We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
@@
 One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area.
-If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
-This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.
+If, during repeat visits, the population is assumed to be closed (no immigration, emigration, reproduction or mortality; i.e., static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability.
+This detection probability can be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.
 
@@
 In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}.
-A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$.
+A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \ Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$.
 Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates.
@@
 There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.
-The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a Frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.
-The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} syntax.
-Models are analyzed using a Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach.
+The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.
+The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} modeling syntax.
+Models are analyzed using Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach.
 The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.
@@
 Overdispersion was generated and modeled using a Negative Binomial distribution for the count component of the model.
-The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter).
+The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.yr = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter).
 All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.
@@
 +                       lam.b2 = -3.0,  # slope for log lambda cov 2
-+                       lam.b4 = 1.0,   # slope for log lambda year
++                       lam.yr = 1.0,   # slope for log lambda year
 +                       p.b0 = 1.0,     # intercept for logit p
@@
 +  # fill abundance array
-+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
++  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.yr*yr)
 +  for(i in 1:n.sites){
@@
   \caption{Differences between the estimated posterior mean values and the true input parameter values as a function of the slope coefficient (effect size) for the site-survey-year covariate, x.p.3.
-Black circles and lines are output from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.
+Black circles and lines are from the model with the site-survey-year covariate, x.p.3, and gray circles and lines are from the model with an averaged site-year covariate, x.p.3.mean.
 Parameter name and true input value is given in the strip across the top of each panel.}
@@
 
-The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
+The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from the \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
 \ref{fig:fig3}).
@@
 \section[Discussion]{Discussion}
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.
+The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.
 
@@
 \end{eqnarray}
-The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
+The log-likelihood can then be evaluated using the following simple \proglang{R}-code.
 \begin{CodeInput}

### 2017-04-25 meehan_et_al_inla_nmix_jss/article.tex
Small manuscript fix

@@
 On the right side of the formula is a 1 to specify a global intercept for $p$ and the two covariates for $p$.
-Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}.
+Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017}.
 The second argument describes the data, provided here as a list that corresponds with the model formula.
@@
 Several other characteristics of the analysis can be modified in the call to inla().
-See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.
+See \cite{Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for details.
 
@@
 \section[Discussion]{Discussion}
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.
+The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.
 

### 2017-04-26 meehan_et_al_inla_nmix_jss/meehan_et_al_inla_nmix_jss.R
fixed reference

@@ -1,6 +1,6 @@
 # code for Meehan, Michel, and Rue #############################################
 # demonstration of R-INLA for analyzing N-mixture models
-# 04/19/2017
+# 04/26/2017
 # ##############################################################################
 
 
@@ -8,9 +8,8 @@
 
 
 # directories, libraries, and helpers ##########################################
-setwd("~/GitHub/Quantitative_Metrics/INLA_NMix")
 library(runjags)
-library(INLA)
+library(INLA) # may require testing version
 library(unmarked)
 library(ggplot2)
 # multiplot function
@@ -677,10 +676,5 @@ dev.off()
 
 
 
-
-
-# wrap up ######################################################################
-save.image("~/GitHub/Quantitative_Metrics/INLA_NMix/meehan_et_al_inla_nmix.RData")
-# ##############################################################################
 
 

### 2017-05-03 unbranded_manuscript/meehan_et_al_inla_nmix_arxiv.tex
added unbranded manuscript to repo

--- 2017-04-25 meehan_et_al_inla_nmix_jss/article.tex
@@
 %% set up %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[article]{jss}
+\documentclass{article}
 \usepackage{amstext}
@@
 \usepackage{natbib}
+\bibliographystyle{plainnat}
+\usepackage{authblk}
+\usepackage[margin=28mm]{geometry}
+\usepackage{titling}
+\newcommand{\subtitle}[1]{%
+	\posttitle{%
+		\par\end{center}
+	\begin{center}\large#1\end{center}
+	\vskip4mm}%
+}
+\title{	Estimating Animal Abundance with N-Mixture Models \\ Using the \emph{R-INLA} Package for \emph{R}}
 
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\author{Timothy D. Meehan\\National\\Audubon Society \And
-        Nicole L. Michel\\National\\Audubon Society \And
-        H{\aa}vard Rue\\King Abdulla University\\of Science and Technology}
-\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}
+\subtitle{DRAFT: NOT YET PEER REVIEWED}
+\author{Timothy D. Meehan$^1$, Nicole L. Michel$^2$, H{\aa}vard Rue$^3$}
+\date{
+	$^1$\emph{National Audubon Society, Boulder, CO USA\\}
+	$^2$\emph{National Audubon Society, San Francisco, CA USA\\}
+	$^3$\emph{King Abdulla University of Science and Technology, Thuwal, Saudi Arabia\\}
+}
 
-%% comma-separated
-\Plainauthor{Timothy D. Meehan, Nicole L. Michel, Haavard Rue}
-%% without formatting
-\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
-%% a short title
-\Shorttitle{\pkg{R-INLA} for N-Mixture Models}
+% start   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+\maketitle
 
-%% an abstract and keywords
-\Abstract{
+\begin{abstract}
+	\fontsize{10pt}{11pt}\selectfont
 	Successful management of wildlife populations requires accurate estimates of abundance.
@@
 N-mixture models enable quantification of detection probability and often produce abundance estimates that are less biased.
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches -- \proglang{JAGS} (via the \pkg{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \pkg{unmarked}, which uses Maximum Likelihood and allows frequentist inference.
-We show that \pkg{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \proglang{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\pkg{R-INLA} is 10 times faster than \pkg{unmarked}, 300 times faster than \proglang{JAGS}), and (4) Bayesian inference is preferred.
-}
-\Keywords{abundance, detection, \proglang{JAGS}, N-mixture model, \proglang{R}, \pkg{R-INLA}, \pkg{unmarked}, wildlife}
-%% without formatting
-\Plainkeywords{wildlife, abundance, detection, N-mixture models, R-INLA, R, BUGS}
-
-%% publication information
-%% NOTE: This can be left commented and will be filled out by the technical editor
-%% \Volume{50}
-%% \Issue{9}
-%% \Month{June}
-%% \Year{2012}
-%% \Submitdate{2012-06-04}
-%% \Acceptdate{2012-06-04}
-
-%% The address of (at least) one author should be given
-\Address{
-  Timothy D. Meehan\\
-  National Audubon Society\\
-  Boulder, Colorado USA\\
-  E-mail: \email{tmeehan@audubon.org}\\
-\\
-  Nicole L. Michel\\
-  National Audubon Society\\
-  San Francisco, California USA\\
-  E-mail: \email{nmichel@audubon.org}\\
-  \\
-  H{\aa}vard Rue\\
-  King Abdulla University
-  of Science and Technology\\
-  Thuwal, Saudi Arabia\\
-  E-mail: \email{haavard.rue@kaust.edu.sa }
-}
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% start main document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
+The purpose of this study was to demonstrate the use of the \emph{R-INLA} package to analyze N-mixture models and to compare performance of \emph{R-INLA} to two other common approaches -- \emph{JAGS} (via the \emph{runjags} package), which uses Markov chain Monte Carlo and allows Bayesian inference, and \emph{unmarked}, which uses Maximum Likelihood and allows frequentist inference.
+We show that \emph{R-INLA} is an attractive option for analyzing N-mixture models when (1) familiar model syntax and data format (relative to other \emph{R} packages) are desired, (2) survey level covariates of detection are not essential, (3) fast computing times are necessary (\emph{R-INLA} is 10 times faster than \emph{unmarked}, 300 times faster than \emph{JAGS}), and (4) Bayesian inference is preferred.
+\end{abstract}
 
 %% introduction
+\fontsize{11pt}{12pt}\selectfont
 \section[Introduction]{Introduction}
@@
 However, there are practical issues that sometimes act as barriers to adoption.
-Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
+Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \emph{WinBUGS}, \emph{OpenBUGS}, or \emph{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}.
 These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.
-First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax.
+First, many wildlife biologists are not accustomed to coding statistical models using the \emph{BUGS} modeling syntax.
 While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment.
@@
 There are other tools available for analyzing N-mixture models that alleviate some of these practical issues.
-The \pkg{unmarked} package \citep{Fiske_Chandler_others_2011} for \proglang{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.
-The model coding syntax used in \pkg{unmarked} is a simple extension of the standard \proglang{R} modeling syntax.
+The \emph{unmarked} package \citep{Fiske_Chandler_others_2011} for \emph{R} statistical computing software \citep{R_Core_Team_2016} offers several options for analyzing N-mixture models within a frequentist framework, with the capacity to accommodate overdispersion and dynamic populations.
+The model coding syntax used in \emph{unmarked} is a simple extension of the standard \emph{R} modeling syntax.
 Models are analyzed using Maximum Likelihood (ML), so model analysis is often completed in a fraction of the time taken using an MCMC approach.
-The familiar model syntax and rapid model evaluation of \pkg{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.
+The familiar model syntax and rapid model evaluation of \emph{unmarked} has undoubtedly contributed to the broader adoption of N-mixture models by wildlife biologists.
 However, it comes at a cost -- loss of the intuitive inferential framework associated with Bayesian analysis.
 
-Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \proglang{R}.
-The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}.
+Here we discuss analysis of N-mixture models using the \emph{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \emph{R}.
+The \emph{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}.
 INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC.
-Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}.
-Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low.
-The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.
+Like \emph{unmarked}, the model syntax used in the \emph{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \emph{R}.
+Also, like \emph{unmarked}, the computational cost of analyzing models with \emph{R-INLA} is relatively low.
+The \emph{R-INLA} approach is different from \emph{unmarked} in that inference about model parameters falls within a Bayesian framework.
 
 \subsection[Overall objectives]{Overall objectives}
-The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \pkg{R-INLA} package.
-In the process, we employ simulated and real count datasets, and analyze them using \proglang{JAGS}, via the \pkg{runjags} package \citep{Denwood_2016} for \proglang{R}, the \pkg{unmarked} package, and the \pkg{R-INLA} package.
+The purpose of this manuscript is to demonstrate analysis of N-mixture models using the \emph{R-INLA} package.
+In the process, we employ simulated and real count datasets, and analyze them using \emph{JAGS}, via the \emph{runjags} package \citep{Denwood_2016} for \emph{R}, the \emph{unmarked} package, and the \emph{R-INLA} package.
 In each case, we demonstrate how data are formatted, how models are specified, how model estimates compare to simulation inputs, and how methods compare in terms of computational performance.
-We also explore a limitation of the \pkg{R-INLA} approach related to model specification.
-Specifically, while it is possible to specify survey level covariates for detection using \proglang{JAGS} and \pkg{unmarked}, this is not possible using \pkg{R-INLA}.
+We also explore a limitation of the \emph{R-INLA} approach related to model specification.
+Specifically, while it is possible to specify survey level covariates for detection using \emph{JAGS} and \emph{unmarked}, this is not possible using \emph{R-INLA}.
 Rather, survey level covariates of detection must be averaged to the site or site-year level.
@@
 However, some repeated code, and code related to generating tables and figures, is not shown for brevity.
-All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}.
-See \url{https://r-inla.org} to connect with the community around the development of \pkg{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.
+All code can be accessed via https://github.com/tmeeha/inlaNMix.
+See https://r-inla.org to connect with the community around the development of \emph{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.
 
@@
 
-\begin{CodeInput}
+\begin{verbatim}
 R> data4sim <- function(n.sites = 72,   # number study sites
@@
 R> sim.data <- data4sim() # create datasets
-\end{CodeInput}
+\end{verbatim}
 
 \subsection[Real data]{Real data}
-In addition to simulated data, we also demonstrate the use of \pkg{R-INLA} with a real dataset in Example III.
-This dataset comes from a study by \cite{Kery_Royle_Schmid_2005} and is publicly available as part of the \pkg{unmarked} package.
+In addition to simulated data, we also demonstrate the use of \emph{R-INLA} with a real dataset in Example III.
+This dataset comes from a study by \cite{Kery_Royle_Schmid_2005} and is publicly available as part of the \emph{unmarked} package.
 The dataset includes mallard duck (\emph{Anas platyrhynchos}) counts, conducted at 239 sites on 2 or 3 occasions during the summer of 2002 as part of a Swiss program that monitors breeding bird abundance (Monitoring H\"{a}ufige Brutv\"{o}gel).
@@
 \subsection[Goals]{Goals}
-In Example I, we demonstrate the use of \pkg{R-INLA} and compare use and performance to similar analyses using \proglang{JAGS} and \pkg{unmarked}.
-In this exercise, the functional forms of \proglang{JAGS}, \pkg{unmarked}, and \pkg{R-INLA} models match the data generating process.
+In Example I, we demonstrate the use of \emph{R-INLA} and compare use and performance to similar analyses using \emph{JAGS} and \emph{unmarked}.
+In this exercise, the functional forms of \emph{JAGS}, \emph{unmarked}, and \emph{R-INLA} models match the data generating process.
 Specifically, we used the covariate x.p.3.mean to generate the count matrix y1 and analyzed the data with models that use x.p.3.mean as a covariate.
@@
 
-\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
-We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package.
+\subsection[Analysis with JAGS]{Analysis with \emph{JAGS}}
+We first analyzed the simulated data using \emph{JAGS}, via the \emph{runjags} package.
 In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.
 
-\begin{CodeInput}
+\begin{verbatim}
 R> jags.model.string <- "
@@
 +  "
-\end{CodeInput}
+\end{verbatim}
 
-Next, we defined the parameters to be monitored during the MCMC runs, bundled the data for \proglang{JAGS}, and created a function for drawing random initial values for the model parameters.
+Next, we defined the parameters to be monitored during the MCMC runs, bundled the data for \emph{JAGS}, and created a function for drawing random initial values for the model parameters.
 The initial values for abundance were made to avoid values of "NA" and zero, as these would cause computational problems.
 
-\begin{CodeInput}
+\begin{verbatim}
 R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
@@
 +               overDisEst = runif(1, 0.5, 2.5))
-\end{CodeInput}
+\end{verbatim}
 
 Finally, we set the run parameters and started the MCMC process.
-Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors < 1.05) and reasonably robust posterior distributions (effective sample sizes > 1000).
+Run parameters were chosen such that MCMC diagnostics indicated converged chains (potential scale reduction factors $<$ 1.05) and reasonably robust posterior distributions (effective sample sizes $>$ 1000).
 Note that the recommended number of effective samples for particularly robust inference is closer to 6000 \citep{Gong_Flegal_2016}.
@@
 
-\begin{CodeInput}
+\begin{verbatim}
 R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
@@
 +                       modules = "glm on", method = "parallel")
-\end{CodeInput}
+\end{verbatim}
 
-Mean parameter estimates from the \proglang{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig.
+Mean parameter estimates from the \emph{JAGS} model were reasonably close to, and not significantly different from, the input values used to generate the data (Fig.
 \ref{fig:fig1}).
-The potential scale reduction factor for all variables was < 1.01, and the effective sample size for all variables was > 2409.
+The potential scale reduction factor for all variables was $<$ 1.01, and the effective sample size for all variables was $>$ 2409.
 The simulation ran in parallel on 3 virtual cores, 1 MCMC chain per core, and took approximately 2104 seconds.
@@
   \includegraphics[width=\linewidth]{fig1.png}
-  \caption{Marginal posteriors from \proglang{JAGS} (solid gray lines) and \pkg{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \pkg{unmarked}.
+  \caption{Marginal posteriors from \emph{JAGS} (solid gray lines) and \emph{R-INLA} (dashed black lines), along with Maximum Liklihood estimates (black circles) and 95\% confidence intervals (horizontal black lines) from \emph{unmarked}.
 True input values are represented by vertical black lines.}
@@
 
-\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
-Next, we prepared the simulated data for the \pkg{unmarked} analysis, which involved slight modification of the data in the sim.data object.
+\subsection[Analysis with unmarked]{Analysis with \emph{unmarked}}
+Next, we prepared the simulated data for the \emph{unmarked} analysis, which involved slight modification of the data in the sim.data object.
 Here, the count data was changed from a 3-dimensional $I*J*K$ array to a 2-dimensional $I*K$ row by $J$ column matrix.
@@
 The site-year variable, x.p.3.mean, was transformed from 2-dimensional matrix to a single column vector.
-Reformatted variables were then assembled in an \pkg{unmarked} data structure called an unmarked frame.
+Reformatted variables were then assembled in an \emph{unmarked} data structure called an unmarked frame.
 
-\begin{CodeInput}
+\begin{verbatim}
 R> y.unmk <- sim.data$y1[ , , 1]
@@
 R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
-\end{CodeInput}
+\end{verbatim}
 
@@
 
-\begin{CodeInput}
+\begin{verbatim}
 R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
@@
 +                       data = unmk.data, mixture = "NB")
-\end{CodeInput}
+\end{verbatim}
 
-Maximum Likelihood estimates from the \pkg{unmarked} analysis are also close to, and not significantly different from, input values (Fig.
-\ref{fig:fig1})  The \pkg{unmarked} estimates were produced in approximately 77 seconds.
+Maximum Likelihood estimates from the \emph{unmarked} analysis are also close to, and not significantly different from, input values (Fig.
+\ref{fig:fig1})  The \emph{unmarked} estimates were produced in approximately 77 seconds.
 
-\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
-Next, we prepared data for \pkg{R-INLA}, which uses data formats similar to those used by \pkg{unmarked}.
-The object counts.and.count.covs is an \pkg{R-INLA} object that includes an $I*K$ row by $J$ column matrix of counts.
+\subsection[Analysis with R-INLA]{Analysis with \emph{R-INLA}}
+Next, we prepared data for \emph{R-INLA}, which uses data formats similar to those used by \emph{unmarked}.
+The object counts.and.count.covs is an \emph{R-INLA} object that includes an $I*K$ row by $J$ column matrix of counts.
 Next comes a value of 1, which is turned into an $I*K$ length vector of ones to specify that the model has a global intercept for $\lambda$.
@@
 
-\begin{CodeInput}
+\begin{verbatim}
 
@@
 +                                      x.lam.2.inla, yr.inla)
-\end{CodeInput}
+\end{verbatim}
 
@@
 The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components.
-In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available.
+In this case, the priors are specified similarly to those in the \emph{JAGS} model, but a variety of other options are available.
 Several other characteristics of the analysis can be modified in the call to inla().
@@
 
-\begin{CodeInput}
+\begin{verbatim}
 R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
@@
 +                                                     param = numeric()))))
-\end{CodeInput}
+\end{verbatim}
 
-Mean parameter estimates from the \pkg{R-INLA} model are also close to, and not significantly different from, input values (Fig.
+Mean parameter estimates from the \emph{R-INLA} model are also close to, and not significantly different from, input values (Fig.
 \ref{fig:fig1}).
-Full marginal posterior distributions from \pkg{R-INLA} are shown in Fig.
+Full marginal posterior distributions from \emph{R-INLA} are shown in Fig.
 \ref{fig:fig1}.
@@
 \subsection[Example I summary]{Example I summary}
-Example I demonstrated basic use of \pkg{R-INLA} and highlighted similarities and differences between it and two other commonly used approaches.
-In demonstrating the use of \pkg{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.
-Similarly, model specification uses a straightforward extension of the standard syntax in \proglang{R}, where the count matrix and covariates for $\lambda$ are specified through an \pkg{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula.
-The data format and model specification syntax of \pkg{R-INLA} are not too different from \pkg{unmarked}, which are both considerably different from \proglang{JAGS} and other \proglang{BUGS}-oriented MCMC software.
+Example I demonstrated basic use of \emph{R-INLA} and highlighted similarities and differences between it and two other commonly used approaches.
+In demonstrating the use of \emph{R-INLA}, we show that the input data format is not complicated, and that the formatting process can be accomplished with relatively few lines of code.
+Similarly, model specification uses a straightforward extension of the standard syntax in \emph{R}, where the count matrix and covariates for $\lambda$ are specified through an \emph{R-INLA} object included on the left side of the formula, and covariates for $p$ are specified on the right side of the formula.
+The data format and model specification syntax of \emph{R-INLA} are not too different from \emph{unmarked}, which are both considerably different from \emph{JAGS} and other \emph{BUGS}-oriented MCMC software.
 
-Regarding performance, \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} successfully extracted simulation input values.
+Regarding performance, \emph{R-INLA}, \emph{unmarked}, and \emph{JAGS} successfully extracted simulation input values.
 Fig.
-\ref{fig:fig1} shows marginal posterior distributions produced by \proglang{JAGS} and \pkg{R-INLA}, and estimates and 95\% confidence intervals from \pkg{unmarked}.
+\ref{fig:fig1} shows marginal posterior distributions produced by \emph{JAGS} and \emph{R-INLA}, and estimates and 95\% confidence intervals from \emph{unmarked}.
 These results derive from data from one random manifestation of the input values.
@@
 Fig.
-\ref{fig:fig1} shows that, for similarly specified models, \pkg{R-INLA} (dashed black lines) and \proglang{JAGS} (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.
+\ref{fig:fig1} shows that, for similarly specified models, \emph{R-INLA} (dashed black lines) and \emph{JAGS} (solid gray lines) yielded practically identical marginal posterior distributions for model parameters.
 
-Where \pkg{R-INLA}, \pkg{unmarked}, and \proglang{JAGS} differed was in computing time.
-In this example, \pkg{R-INLA} took 6 seconds, \pkg{unmarked} took 77 seconds, and \proglang{JAGS} took 2104 seconds to produce results.
-Thus, \pkg{R-INLA} was approximately 10 times faster than \pkg{unmarked} and 300 times faster than \proglang{JAGS}.
-This was the case despite the fact that \pkg{unmarked} produces ML estimates and the \proglang{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core.
-If parallel computing had not been used with \proglang{JAGS}, processing the \proglang{JAGS} model would have taken approximately twice as long.
+Where \emph{R-INLA}, \emph{unmarked}, and \emph{JAGS} differed was in computing time.
+In this example, \emph{R-INLA} took 6 seconds, \emph{unmarked} took 77 seconds, and \emph{JAGS} took 2104 seconds to produce results.
+Thus, \emph{R-INLA} was approximately 10 times faster than \emph{unmarked} and 300 times faster than \emph{JAGS}.
+This was the case despite the fact that \emph{unmarked} produces ML estimates and the \emph{JAGS} model was run in parallel with each of three MCMC chains simulated on a separate virtual computing core.
+If parallel computing had not been used with \emph{JAGS}, processing the \emph{JAGS} model would have taken approximately twice as long.
 If MCMC simulations were run until an effective sample size of 6000 was reached, processing times would have doubled again.
 
-In sum, when compared to other tools, \pkg{R-INLA} is relatively easy to implement and produces accurate estimates of Bayesian posteriors very quickly.
+In sum, when compared to other tools, \emph{R-INLA} is relatively easy to implement and produces accurate estimates of Bayesian posteriors very quickly.
 Its utility depends on the degree to which the data generating process can be captured accurately in model specification.
-However, as mentioned above, certain N-mixture models can not be specified using \pkg{R-INLA}.
+However, as mentioned above, certain N-mixture models can not be specified using \emph{R-INLA}.
 For the data in Example I, the count matrix was produced using a detection covariate that was averaged to the site-year level.
@@
 Specifically, we conducted a Monte Carlo simulation where, for each iteration, the count matrix for the analysis, y2, was generated with the data generating function described in 1.3, using the site-survey-year covariate x.p.3.
-The count data were then analyzed with two \proglang{JAGS} models.
+The count data were then analyzed with two \emph{JAGS} models.
 One model incorporated an averaged site-survey x.p.3.mean as a covariate, such as in 2.1.
@@
 
-\subsection[Analysis with JAGS]{Analysis with \proglang{JAGS}}
-Given the long processing time associated with \proglang{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation.
+\subsection[Analysis with JAGS]{Analysis with \emph{JAGS}}
+Given the long processing time associated with \emph{JAGS} models, we only ran 3000 iterations (no thinning, after 1000 adaptive and 1000 burn-in iterations) during each simulation.
 This number is not sufficient for drawing inference from marginal posteriors, but was sufficient for looking at qualitative patterns in posterior means.
@@
 \subsection[Goals]{Goals}
-In Example III, we explore the performance of \pkg{R-INLA} using real data -- a publicly available dataset on mallard duck abundance in Switzerland in 2002.
-By employing real data, we hoped to evaluate (1) the performance of \pkg{R-INLA} using data that were not predictable by design and (2) the practical outcome of not being able to specify site-survey covariates in \pkg{R-INLA}.
-The dataset is available as a demonstration dataset in \pkg{unmarked}, so we compared the performance of \pkg{R-INLA} with \pkg{unmarked}, using the demonstration model and analysis settings, in Example III.
+In Example III, we explore the performance of \emph{R-INLA} using real data -- a publicly available dataset on mallard duck abundance in Switzerland in 2002.
+By employing real data, we hoped to evaluate (1) the performance of \emph{R-INLA} using data that were not predictable by design and (2) the practical outcome of not being able to specify site-survey covariates in \emph{R-INLA}.
+The dataset is available as a demonstration dataset in \emph{unmarked}, so we compared the performance of \emph{R-INLA} with \emph{unmarked}, using the demonstration model and analysis settings, in Example III.
 
-\subsection[Analysis with unmarked]{Analysis with \pkg{unmarked}}
-The \pkg{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection.
+\subsection[Analysis with unmarked]{Analysis with \emph{unmarked}}
+The \emph{unmarked} model was specified with linear effects of transect length (length), elevation (elev), and forest cover (forest) on abundance, and a linear effect and quadratic effect of sampling intensity (ivel) and sampling date (date), respectively, on detection.
 Both sampling intensity and sampling date were site-survey, or survey level covariates.
 
-\begin{CodeInput}
+\begin{verbatim}
 R> data("mallard") # get mallard data from unmarked package
@@
 +                               mixture = "NB", mallardUMF, K=30) 
-\end{CodeInput}
+\end{verbatim}
 
-\subsection[Analysis with R-INLA]{Analysis with \pkg{R-INLA}}
-The \pkg{R-INLA} model used a similar model structure.
-The only difference was that the \pkg{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead.
+\subsection[Analysis with R-INLA]{Analysis with \emph{R-INLA}}
+The \emph{R-INLA} model used a similar model structure.
+The only difference was that the \emph{R-INLA} model was not able to process site-survey covariates for detection, so averaged site level covariates were used, instead.
 These averages were attained using the rowMeans() function.
 
-\begin{CodeInput}
+\begin{verbatim}
 R> mallard.y.inla <- mallard.y # format inla data
@@
 +                                                  param = numeric()))))
-\end{CodeInput}
+\end{verbatim}
 
-The 95\% credible intervals from the \pkg{R-INLA} model overlapped broadly with the 95\% confidence intervals from the \pkg{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
+The 95\% credible intervals from the \emph{R-INLA} model overlapped broadly with the 95\% confidence intervals from the \emph{unmarked} model, so parameter estimates were not significantly different from one another (Fig.
 \ref{fig:fig3}).
@@
 Using both techniques, detection decreased linearly as the season progressed, and abundance decreased as linear functions of increasing forest cover and elevation gain.
-Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \pkg{unmarked} and site-averaged detection covariates were used for \pkg{R-INLA}.
+Parameter estimates and biological conclusions were similar despite the fact that site-survey detection covariates were used for \emph{unmarked} and site-averaged detection covariates were used for \emph{R-INLA}.
 Note that both approaches estimated moderate effects of detection covariates which, according to the results in Example II, would indicate that other parameter estimates were not substantially biased.
@@
   \includegraphics[width=\linewidth]{fig3.png}
-  \caption{Estimates and 95\% confidence intervals from \pkg{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \pkg{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.
-The \pkg{unmarked} model included site-survey covariates for survey effort and survey date, while the \pkg{R-INLA} model included site-averaged versions.
+  \caption{Estimates and 95\% confidence intervals from \emph{unmarked} (gray circles and lines) and posterior means and 95\% credible intervals from \emph{R-INLA} (black lines) from an N-mixture model of mallard duck abundance.
+The \emph{unmarked} model included site-survey covariates for survey effort and survey date, while the \emph{R-INLA} model included site-averaged versions.
 A value of zero is depicted by the vertical gray line.}
@@
 \section[Discussion]{Discussion}
-The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.
+The purpose of this study was to demonstrate the use of the \emph{R-INLA} package \citep{Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} to analyze N-mixture models and to compare performance of \emph{R-INLA} to two other common approaches, \emph{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \emph{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \emph{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows frequentist inference.
 
-Results showed that \pkg{R-INLA} can be a complementary tool in the wildlife biologist's analytical tool kit.
-Strengths of \pkg{R-INLA} include Bayesian inference, based on highly accurate approximations of posterior distributions, which are derived over 300 times faster than MCMC methods, where models are specified using a syntax that should be familiar to R users, and where data are formatted in a straightforward way with relatively few lines of code.
-The straightforward model syntax and data format could help remove  barriers to adoption of N-mixture models for biologists who are not committed to learning the \proglang{BUGS} syntax.
+Results showed that \emph{R-INLA} can be a complementary tool in the wildlife biologist's analytical tool kit.
+Strengths of \emph{R-INLA} include Bayesian inference, based on highly accurate approximations of posterior distributions, which are derived over 300 times faster than MCMC methods, where models are specified using a syntax that should be familiar to R users, and where data are formatted in a straightforward way with relatively few lines of code.
+The straightforward model syntax and data format could help remove  barriers to adoption of N-mixture models for biologists who are not committed to learning the \emph{BUGS} syntax.
 The substantial decrease in computation time should facilitate use of a wider variety of model and variable selection techniques (e.g., cross validation and model averaging), ones that are not commonly used in an MCMC context due to practical issues related to computing time \citep{Kery_Schaub_2011}.
 
-Limitations of \pkg{R-INLA} are mainly related to the more restricted set of N-mixture models that can be specified.
-Of the approaches explored here, \proglang{BUGS}-based approaches allow users ultimate flexibility in specifying models.
-For example, with \proglang{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}.
-In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-Binomial and Negative Binomial-Binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.
-In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig.
+Limitations of \emph{R-INLA} are mainly related to the more restricted set of N-mixture models that can be specified.
+Of the approaches explored here, \emph{BUGS}-based approaches allow users ultimate flexibility in specifying models.
+For example, with \emph{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}.
+In comparison, the current version of \emph{R-INLA} does not handle site-survey covariates, employs only Poisson-Binomial and Negative Binomial-Binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only.
+In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \emph{R-INLA} will not be the appropriate tool (Fig.
 \ref{fig:fig2}).
 
-When compared to \pkg{unmarked}, the \pkg{R-INLA} approach is similar in regards to model syntax and data format.
+When compared to \emph{unmarked}, the \emph{R-INLA} approach is similar in regards to model syntax and data format.
 The approaches are also similar in that both yield results much faster than MCMC.
-The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, likely due to the different approach used to compute model likelihoods (see Appendix A).
-They also differ in that \pkg{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
+The two approaches differ in that \emph{R-INLA} is roughly 10 times faster than \emph{unmarked}, likely due to the different approach used to compute model likelihoods (see Appendix A).
+They also differ in that \emph{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.
 
-In conclusion, \proglang{JAGS} (and \proglang{WinBUGS}, \proglang{OpenBUGS}), \pkg{unmarked}, and \pkg{R-INLA} all allow users to analyze N-mixture models for estimating wildlife abundance while accounting for imperfect detection.
+In conclusion, \emph{JAGS} (and \emph{WinBUGS}, \emph{OpenBUGS}), \emph{unmarked}, and \emph{R-INLA} all allow users to analyze N-mixture models for estimating wildlife abundance while accounting for imperfect detection.
 Each method has its strengths and limitations.
-\pkg{R-INLA} appears to be an attractive option when survey level covariates are not essential, familiar model syntax and data format are desired, fast computing time is needed, and Bayesian inference is preferred.
-
-%% acknowledgements
-\section*{Acknowledgments}
-We thank Jerry Cole and Curtis Burkhalter for commenting on previous drafts of this manuscript.
+\emph{R-INLA} appears to be an attractive option when survey level covariates are not essential, familiar model syntax and data format are desired, fast computing time is needed, and Bayesian inference is preferred.
 
@@
 \end{displaymath}
-where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{Binomial}(y; n, p)$ is the density for the Binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
+where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with a mean $\lambda$, $\lambda^{n}\frac{\exp(-\lambda)}{n!}$, and $\text{Binomial}(y; n, p)$ is the density for the Binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$.
 Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we can easily evaluate it using a recursive algorithm that is both faster and more numerical stable.
@@
 \end{eqnarray}
-The log-likelihood can then be evaluated using the following simple \proglang{R}-code.
-\begin{CodeInput}
+The log-likelihood can then be evaluated using the following simple \emph{R}-code.
+\begin{verbatim}
 	R> fac <- 1; ff <- lambda * (1-p)
@@
 	+           log(fac)
-\end{CodeInput}
+\end{verbatim}
 Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small.
-Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
+Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
 

### 2017-05-03 unbranded_manuscript/meehan_et_al_inla_nmix_jss.R

@@ -1,0 +1,686 @@
+# code for Meehan, Michel, and Rue #############################################
+# demonstration of R-INLA for analyzing N-mixture models
+# 04/19/2017
+# ##############################################################################
+
+
+
+
+
+# directories, libraries, and helpers ##########################################
+setwd("~/GitHub/Quantitative_Metrics/INLA_NMix")
+library(runjags)
+library(INLA)
+library(unmarked)
+library(ggplot2)
+# multiplot function
+multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
+  library(grid)
+  plots <- c(list(...), plotlist)
+  numPlots = length(plots)
+  if (is.null(layout)) {
+    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
+                     ncol = cols, nrow = ceiling(numPlots/cols))
+  }
+  if (numPlots==1) {
+    print(plots[[1]])
+  } else {
+    grid.newpage()
+    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
+    for (i in 1:numPlots) {
+      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
+      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
+                                      layout.pos.col = matchidx$col))
+    }
+  }
+}
+# plotting theme
+theme_acbs <- function (base_size = 11, base_family = "") {
+  theme_grey(base_size = base_size, base_family = base_family) %+replace%
+    theme(panel.background = element_rect(fill = "white", colour = NA),
+          panel.border = element_rect(fill = NA, colour = "grey20"),
+          panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          axis.text.x = element_text(size = rel(0.9), angle = 0),
+          axis.text.y = element_text(size = rel(0.9), angle = 0),
+          strip.background = element_rect(fill = "grey85", colour = "grey20"),
+          legend.key = element_rect(fill = "white", colour = NA),
+          complete = TRUE)
+}
+# data generating function
[…]

### 2017-05-09 meehan_et_al_inla_nmix_jss/article.tex
more tweaks
Minor edit of the R code in the Appendix

@@
 \usepackage{natbib}
+\usepackage{hyperref}
 
@@
 R> fac <- 1; ff <- lambda * (1-p)
-R> for(i in (n.max - y):1){
-+    fac <- 1 + fac * ff / i
-+  } 
-R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
-+           log(fac)
+R> for(i in (n.max - y):1) fac <- 1 + fac * ff / i
+R> log.L <- dpois(y, lambda, log = TRUE) +
++           dbinom(y, y, p, log = TRUE) + log(fac)
 \end{CodeInput}
